---
title: "Stat 135 Discussion Section 1"
subtitle: "R Tutorial: Estimation, Random Sampling, and Parametric Models"
output:
  pdf_document:
    toc: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)

# This seed makes all simulated results reproducible.
set.seed(13501)
```

# How to use this tutorial

This tutorial combines statistical explanations with executable R code.
Run the chunks in order because later chunks use objects created
earlier. In RStudio, you can run a chunk using the green arrow at its
upper-right corner or knit the entire document using **Knit**.

The statistical code uses base R. The `knitr` package is used only to
format tables and render the document.

By the end of the tutorial, you should be able to:

1.  explain why the mean of zero-one data equals the sample proportion;
2.  draw a simple random sample without replacement;
3.  use R's `d*`, `p*`, `q*`, and `r*` probability functions; and
4.  compare simulated data with a theoretical density or probability
    mass function.

# Estimation with binary data

## The mean of binary observations

Suppose a researcher wants to estimate the proportion of voters in a
city who support a new public transportation proposal.

A random sample of $n=20$ voters is asked:

“Do you support the proposal?”

For voter $i$ in the sample, define

$$
X_i=
\begin{cases}
1, &\text{if observation $i$ is a success}\\
0, &\text{otherwise}
\end{cases}.
$$

Assume

$$ 
X_i \sim \operatorname{Bernoulli}(\pi), 
$$

where $\pi$ is the true proportion of all voters in the city who would
vote Yes, and $X_i$s are independent of each other.

Assume there are $k$ Yes votes out of $n$ voters in the sample, then the
sample proportion is

$$
\hat \pi = \frac{k}{n}.
$$

$k=\sum_{i=1}^n X_i$ because each Yes vote contributes 1 to the sum and
each No vote contributes 0. Then the sample mean is

$$ \bar X=\frac{1}{n}\sum_{i=1}^n X_i=\frac{k}{n}=\widehat \pi. $$

Thus, when binary outcomes are coded as 0 and 1, the sample mean is
exactly the sample proportion.

Suppose the observed responses are

$(1,1,0,1,0,1,1,1,0,1,1,1,0,1,0,1,1,1,0,1)$.

There are $k=14$ Yes votes out of $n=20$ voters in the sample.
Therefore, the sample proportion is

$$
\hat \pi = \frac{k}{n} = \frac{14}{20} = 0.7.
$$

So 70% of the sampled voters support the proposal, and we use the sample
proportion

$\hat \pi = 0.7$

to estimate the unknown proportion of all voters in the city who support
the proposal (population proportion).

```{r binary-data}
x <- c(1,1,0,1,0,1,1,1,0,1,1,1,0,1,0,1,1,1,0,1)

n <- length(x)
pi_hat <- mean(x)

x
table(x)
n
pi_hat
```

## Variance of binary observations

Suppose now we want to estimate the population variance. Consider a
variance estimator:

$$
\widehat\sigma^2
=\frac{1}{n}\sum_{i=1}^n(X_i-\bar X)^2.
$$

For binary data, $X_i^2=X_i$.

$$
\widehat\sigma^2=\widehat \pi(1-\widehat \pi).
$$

We can verify both the binary identity and the variance formula
directly.

```{r binary-variance}
# For every zero-one observation, x^2 equals x.
all(x^2 == x)

# Variance estimate using divisor n.
sigma2_hat_direct <- mean((x - mean(x))^2)

# Bernoulli plug-in formula.
sigma2_hat_binary <- pi_hat * (1 - pi_hat)

c(
  direct_calculation = sigma2_hat_direct,
  bernoulli_shortcut = sigma2_hat_binary
)
```

R's `var()` function uses divisor $n-1$:

$$
S^2=\frac{1}{n-1}\sum_{i=1}^n(X_i-\bar X)^2
=\frac{n}{n-1}\widehat\sigma^2.
$$

```{r sample-variance}
S2 <- var(x)
S2_from_sigma2_hat <- n / (n - 1) * sigma2_hat_direct

c(
  R_var = S2,
  converted_from_sigma2_hat = S2_from_sigma2_hat
)
```

Under iid sampling, $S^2$ is unbiased for the population variance.

# Simple random sampling without replacement

## Define the finite population

Consider the finite population

$$
1,\ 3,\ 3,\ 5,\ 8.
$$

The two observations equal to 3 are distinct population units even
though their measured values are identical.

```{r define-population}
population <- c(1, 3, 3, 5, 8)
N <- length(population)
sample_size <- 3

population_mean = mean(population)
population_mean
```

## Draw one simple random sample

In a simple random sample of size $n$ without replacement, every subset
of $n$ distinct population units has the same probability.

```{r one-srs}
one_sample <- sample(
  population,
  size = sample_size,
  replace = FALSE
)

one_sample
c(
  sample_mean = mean(one_sample),
  sample_variance = var(one_sample)  # var() uses divisor $n-1$
)
```

Running the chunk again generally produces a different sample. The
argument `replace = FALSE` prevents a population unit from being
selected more than once.

> **Think:** What would change if `replace = TRUE`? Could the same
> population unit appear more than once?

## Enumerate all possible samples

There are

$$
\binom{5}{3}=10
$$

equally likely unordered samples. We enumerate indices rather than
values so that the two population units with value 3 remain
distinguishable.

```{r enumerate-samples}
index_samples <- combn(seq_along(population), sample_size)

sample_values <- apply(
  index_samples,
  MARGIN = 2,
  function(ind) population[ind]
)

sample_means <- colMeans(sample_values)
sample_medians <- apply(sample_values, MARGIN = 2, median)

sample_table <- data.frame(
  sample = apply(sample_values, 2, paste, collapse = ", "),
  mean = sample_means,
  median = sample_medians
)

sample_table
```

## Sampling distributions

Because all ten samples are equally likely, each row of the preceding
table has probability (1/10). We can aggregate rows that produce the
same statistic.

```{r exact-sampling-distributions}

mean_distribution <- as.data.frame(
  prop.table(table(sample_means))
)

mean_distribution$sample_means <-
  as.numeric(as.character(mean_distribution$sample_means))

knitr::kable(
  as.data.frame(mean_distribution),
  col.names = c("Sample mean", "Probability"),
  digits = 4,
  caption = "Sampling distribution of the sample mean"
)


median_distribution <- as.data.frame(
  prop.table(table(sample_medians))
)
median_distribution$sample_medians <-
  as.numeric(as.character(median_distribution$sample_medians))


knitr::kable(
  as.data.frame(median_distribution),
  col.names = c("Sample median", "Probability"),
  digits = 4,
  caption = "Sampling distribution of the sample median"
)
```

The exact sampling distribution of the sample mean is

$$
\begin{array}{c|cccccc}
\bar X & 7/3 & 3 & 11/3 & 4 & 14/3 & 16/3\\ \hline
\Pr(\bar X=x) & 1/10 & 2/10 & 1/10 & 2/10 & 2/10 & 2/10.
\end{array}
$$

The sample median equals 3 with probability (7/10) and 5 with
probability (3/10).

## Expected value and standard error of the sample mean

The expected value is the probability-weighted average of the possible
values.

```{r sample-mean-se}
mean_of_sample_means <- mean(sample_means)

# These ten values are the complete distribution, so use divisor 10.
var_of_sample_means <- mean(
  (sample_means - mean_of_sample_means)^2
)
se_of_sample_means <- sqrt(var_of_sample_means)

c(
  E_Xbar = mean_of_sample_means,
  Var_Xbar = var_of_sample_means,
  SE_Xbar = se_of_sample_means
)
```

Thus,

$$
\operatorname{E}(\bar X)=4,\qquad
\operatorname{Var}(\bar X)=\frac{14}{15},\qquad
\operatorname{SE}(\bar X)=\sqrt{\frac{14}{15}}\approx0.9661.
$$

Do not use `sd(sample_means)` for the exact standard error of sample
means. The ten values are the entire distribution, whereas `sd()` uses
divisor (10-1), treating them as a sample from a larger population.

## Simulation

When exact enumeration is impractical, repeated random sampling can
approximate the sampling distribution.

```{r simulate-sample-means}
set.seed(13501)
B <- 10000

sim_sample_means <- replicate(
  B,
  mean(sample(population, size = 3, replace = FALSE))
)

c(
  mean_of_sim_sample_means = mean(sim_sample_means),
  sd_of_sim_sample_means = sd(sim_sample_means)
)
```

Each value in `sim_sample_means` is one statistic computed from an
entire random sample. Their histogram approximates the sampling
distribution of $\bar X$.

## Population distribution versus sampling distribution

The first plot contains individual population values. The second
contains statistics—one sample mean for each simulated sample.

```{r compare-histograms, fig.show='hold', out.width='49%'}
par(mfrow = c(1, 2))
hist(population, main = "Population values", xlab = "Value")
abline(v = mean(population), col = "firebrick", lwd = 3)
hist(sim_sample_means, probability = TRUE, main = "Simulated sampling distribution \nof the sample mean", xlab = "Sample mean")
abline(v = mean(population), col = "firebrick", lwd = 3)
par(mfrow = c(1, 1))
```

> **Think:** What does one value in the left histogram represent? What
> does one value in the right histogram represent?

## Binary population example

Now consider a finite population containing 40 ones and 60 zeros. Its
population proportion is 0.40. We randomly sample 20 units without
replacement.

```{r binary-population}
binary_population <- c(rep(1, 40), rep(0, 60))

p_population = mean(binary_population)

N_binary <- length(binary_population)
n_binary <- 20

one_binary_sample <- sample(
  binary_population,
  size = n_binary,
  replace = FALSE
)

one_binary_sample
c(
  population_proportion = mean(binary_population),
  sample_proportion = mean(one_binary_sample),
  population_variance = p_population * (1 - p_population),
  sample_variance = var(one_binary_sample)  # var() uses divisor $n-1$
)
```

```{r plot-binary-sampling-distribution}
B_binary <- 10000

sim_sample_proportions <- replicate(
  B_binary,
  mean(sample(
    binary_population,
    size = n_binary,
    replace = FALSE
  ))
)
hist(
  sim_sample_proportions,
  probability = TRUE,
  main = "Simulated sampling distribution of sample proportion",
  xlab = "Sample proportion"
)
abline(v = p_population, col = "firebrick", lwd = 3)
```

# Parametric probability models in R

## The `d*`, `p*`, `q*`, and `r*` convention

For a distribution named `xxx`, R usually provides four related
functions:

| Function       | Meaning                                  |
|:---------------|:-----------------------------------------|
| `dxxx(x, ...)` | Density value or probability mass at `x` |
| `pxxx(q, ...)` | Cumulative probability $\Pr(X\le q)$     |
| `qxxx(p, ...)` | The $p$-quantile, or inverse CDF         |
| `rxxx(n, ...)` | `n` random draws                         |

For a **continuous** distribution, `dxxx(x)` is a density value, not
$\Pr(X=x)$; the probability at one exact point is zero, $\Pr(X=x) = 0$.
For a **discrete** distribution, `dxxx(x)` is the probablity mass at
$x$, $\Pr(X=x)$.

## Normal distribution

If $X\sim\mathcal N(\mu,\sigma^2)$, then $X$ is continuous random
variable with mean $\mu$ and variance $\sigma^2$.

```{r normal-probabilities}
c(
  P_Z_le_1.96 = pnorm(1.96),
  q_0.975 = qnorm(0.975)
)
```

$\Phi(1.96)\approx0.975$ and $\Phi^{-1}(0.975)\approx1.96$ .

```{r normal-simulation}
normal_draws <- rnorm(10000, mean = 10, sd = 2)

hist(
  normal_draws,
  probability = TRUE,
  main = "Normal simulation with density overlay",
  xlab = "x",
  col = "gray85",
  border = "white"
)

curve(
  dnorm(x, mean = 10, sd = 2),
  from = 2,
  to = 18,
  add = TRUE,
  col = "firebrick",
  lwd = 3
)
```

The histogram must use `probability = TRUE` so that its vertical scale
matches the theoretical density.

> **Think:** Is `dnorm(10, mean = 10, sd = 2)` a probability? What is
> $\Pr(X=10)$ for this continuous distribution?

## Bernoulli distribution

A Bernoulli variable takes values 0 and 1. For
$X\sim \text{Bernoulli}(p)$,

$$
\operatorname{E}(X)=p,\qquad
\operatorname{Var}(X)=p(1-p).
$$

Base R represents $\text{Bernoulli}(p)$ as a Binomial distribution with
`size = 1`.

```{r bernoulli-model}
p <- 0.30

dbinom(0:1, size = 1, prob = p)

c(
  theoretical_mean = p,
  theoretical_variance = p * (1 - p)
)

bernoulli_draws <- rbinom(10000, size = 1, prob = p)

c(
  simulated_mean = mean(bernoulli_draws),
  simulated_variance = var(bernoulli_draws)
)
```
