---
title: "Maximum Likelihood Estimation: Poisson and Exponential"
author: "STAT 135 Lab 2"
date: "Fall 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. The Poisson Model (Asbestos Counts)

**Step 1: Decide on Parametric Model**
We have a dataset of asbestos fiber counts from 23 grid squares, collected during a study by the National Institute of Science and Technology. We must *choose* how to model this data. Because it represents counts of events in a fixed spatial area, we decide to model the data $X_1, \dots, X_n$ as \textit{i.i.d.} $\text{Poisson}(\lambda)$. We assume independence between individual grid square counts and a constant average concentration rate across all observed grid squares.

**Question:** A key property of the Poisson distribution is that its mean equals its variance ($\mathbb{E}[X] = \text{Var}(X) = \lambda$). What alternative distribution might you consider if the observed variance across the grid squares turns out to be substantially larger than the sample mean (overdispersion)?

**Question:** Would modeling the data using a binomial distribution make sense in this setting?

```{r poisson-data}
# Data from Steel et al. 1980 (National Institute of Science and Technology)
asbestos_counts <- c(31, 29, 19, 18, 31, 28, 
                     34, 27, 34, 30, 16, 18, 
                     26, 27, 27, 18, 24, 22, 
                     28, 24, 21, 17, 24)
n_pois <- length(asbestos_counts)
```

**Step 2: Calculate Likelihood & Log-Likelihood**

The Probability Mass Function (PMF) for an individual observation $x_i \in \{0, 1, 2, \dots\}$ is:
$$f(x_i; \lambda) = P(X_i = x_i) = \frac{\lambda^{x_i} e^{-\lambda}}{x_i!}$$

Under the assumption that our observations $X_1, \dots, X_n$ are independent and identically distributed, the **likelihood function** is the joint probability evaluated at the observed data vector $(x_1, \dots, x_n)$ as a function of the unknown parameter $\lambda$:
$$L(\lambda | x_1 \dots x_n) = \prod_{i=1}^n f(x_i; \lambda) = \prod_{i=1}^n \frac{\lambda^{x_i} e^{-\lambda}}{x_i!} = \frac{\lambda^{\sum_{i=1}^n x_i} e^{-n\lambda}}{\prod_{i=1}^n x_i!}$$
Here, we plot the Probability Mass Function (PMF) side-by-side with the Likelihood function. Notice the axes:

* **PMF**: Function of the data $x$, with the parameter $\lambda$ fixed.
* **Likelihood**: Function of the parameter $\lambda$, with the data $x$ fixed.

```{r poisson-pmf-vs-lik, fig.width=10, fig.height=4}
par(mfrow=c(1,2))

# Plotting the PMF (Fixed Lambda, Varying X)
lambda_fixed <- mean(asbestos_counts) # Using sample mean for demonstration
x_vals <- 5:45
pmf_vals <- dpois(x_vals, lambda_fixed)

plot(x_vals, pmf_vals, type="h", lwd=3, col="blue",
     main=paste("PMF: Poisson (lambda =", round(lambda_fixed, 2), ")"),
     xlab="x (Data)", ylab="P(X = x)")
points(x_vals, pmf_vals, pch=16, col="blue")

# Calculate the likelihood as a function of lambda
lik_pois <- function(lambda) {
  prod(dpois(asbestos_counts, lambda))
}
# Plotting the Likelihood (Fixed Data, Varying Lambda)
lambda_grid <- seq(15, 35, by=0.1)
lik_vals <- sapply(lambda_grid, lik_pois)

plot(lambda_grid, lik_vals, type="l", lwd=2, col="red",
     main="Likelihood Function (Given Data)",
     xlab=expression(lambda~" (Parameter)"), ylab="L("~lambda~" | Data)")
```

Now, let's look at the Likelihood vs. the Log-Likelihood. Maximizing one maximizes the other!

```{r poisson-lik-vs-loglik, fig.width=10, fig.height=4}
par(mfrow=c(1,2))

log_lik_pois <- function(lambda) {
  sum(dpois(asbestos_counts, lambda, log=TRUE))
}
log_lik_vals <- sapply(lambda_grid, log_lik_pois)

# Likelihood
plot(lambda_grid, lik_vals, type="l", lwd=2, col="red",
     main="Likelihood Function",
     xlab=expression(lambda), ylab="L("~lambda~" | Data)")
abline(v = mean(asbestos_counts), lty=2)

# Log-Likelihood
plot(lambda_grid, log_lik_vals, type="l", lwd=2, col="darkgreen",
     main="Log-Likelihood Function",
     xlab=expression(lambda), ylab="l("~lambda~" | Data)")
abline(v = mean(asbestos_counts), lty=2)
```

**Exercise:** Derive the log-likelihood function, $\ell(\lambda) = \log L(\lambda \mid x_1, \dots, x_n)$, for an i.i.d. $\text{Poisson}(\lambda)$ sample.

**Note:** Throughout this lab, "log" always refers to the natural logarithm, base $e$ or $\ln$.

**Step 3 & 4: Find Maximum & Plug in Data**

Differentiating the log-likelihood and setting it to zero shows that the maximum occurs at the sample mean.

Our MLE estimate is $\widehat{\lambda} = \bar{X}$.

```{r poisson-estimate}
lambda_hat <- mean(asbestos_counts)
cat("The MLE estimate for lambda is:", lambda_hat, "\n")
```

---

## 2. The Exponential Model (Simulated Data & Parameter Analysis)

The Probability Density Function (PDF) for an individual continuous observation $x_i > 0$ is:
$$f(x_i; \theta) = \theta e^{-\theta x_i}$$

Under the assumption of an independent and identically distributed sample, the **likelihood function** is the joint density evaluated at the observed data vector $(x_1, \dots, x_n)$ as a function of the unknown rate parameter $\theta$:
$$L(\theta | X_1 \dots X_n) = \prod_{i=1}^n f(x_i; \theta) = \prod_{i=1}^n \left( \theta e^{-\theta x_i} \right) = \theta^n \exp\left(-\theta \sum_{i=1}^n x_i\right)$$

Taking the natural logarithm gives the **log-likelihood function**:
$$\ell(\theta | X_1 \dots X_n) = \log L(\theta) = n\log(\theta) - \theta \sum_{i=1}^n x_i$$

Here, we plot the Probability Density Function (PDF) side-by-side with the Likelihood function. Notice the axes:

* **PDF**: Function of the data $x$, with the parameter $\theta$ fixed.
* **Likelihood**: Function of the parameter $\theta$, with the data $x$ fixed.

```{r exp-pdf-vs-lik, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

# Using simulated exponential data for demonstration
set.seed(135)
sample_exp_demo <- rexp(25, rate = 2.0)
theta_fixed <- 1 / mean(sample_exp_demo)

# Plotting the PDF (Fixed Theta, Varying X)
x_vals_exp <- seq(0, 3, length.out = 200)
pdf_vals <- dexp(x_vals_exp, rate = theta_fixed)

plot(x_vals_exp, pdf_vals, type = "l", lwd = 2.5, col = "blue",
     main = paste("PDF: Exponential (theta =", round(theta_fixed, 2), ")"),
     xlab = "x (Data)", ylab = "f(x | theta)")

# Plotting the Likelihood (Fixed Data, Varying Theta)
lik_exp <- function(theta) {
  prod(dexp(sample_exp_demo, rate = theta))
}
theta_grid <- seq(0.5, 4.5, by = 0.01)
lik_vals_exp <- sapply(theta_grid, lik_exp)

plot(theta_grid, lik_vals_exp, type = "l", lwd = 2.5, col = "red",
     main = "Likelihood Function (Given Data)",
     xlab = expression(theta ~ " (Parameter)"), ylab = "L(" ~ theta ~ " | Data)")
```

Next, we simulate exponential data to examine how the likelihood function shifts both when the true parameter ($\theta$) changes and when the sample size changes.  

### Part A: Effect of Different True Parameters ($\theta$)

Let's see how the likelihood function shifts position when data is generated from different true parameters using a fixed sample size ($n = 30$).

```{r exp-diff-params}
set.seed(135)
n_fixed <- 30
true_theta_1 <- 1.0
true_theta_2 <- 2.5

# Note: In R's dexp/rexp, the 'rate' argument corresponds to our parameter theta
data_theta1 <- rexp(n_fixed, rate = true_theta_1)
data_theta2 <- rexp(n_fixed, rate = true_theta_2)

log_lik_exp <- function(theta, data) {
  sum(dexp(data, rate=theta, log=TRUE))
}

theta_grid_exp <- seq(0.1, 4, by=0.02)

ll_theta1 <- sapply(theta_grid_exp, log_lik_exp, data=data_theta1)
ll_theta2 <- sapply(theta_grid_exp, log_lik_exp, data=data_theta2)

plot(theta_grid_exp, ll_theta1 - max(ll_theta1), type="l", col="purple", lwd=2,
     main="Likelihood Function Shifts with Different True Parameters",
     xlab=expression(theta), ylab="Scaled Log-Likelihood")
lines(theta_grid_exp, ll_theta2 - max(ll_theta2), col="orange", lwd=2)
abline(v = true_theta_1, lty=2, lwd=2, col="purple")
abline(v = true_theta_2, lty=2, lwd=2, col="orange")
legend("topright", legend=c("True Theta = 1.0", "True Theta = 2.5"),
       col=c("purple", "orange"), lwd=2)
```

### Part B: Effect of Different Sample Sizes

Next, holding the true parameter constant ($\theta = 2$), we look at how larger sample sizes cause the likelihood to become more concentrated near its maximum. 

```{r exp-sim-data}
set.seed(135)
true_theta <- 2

# Simulating two datasets of different sizes
n_small <- 10
n_large <- 1000

data_small <- rexp(n_small, rate = true_theta)
data_large <- rexp(n_large, rate = true_theta)
```

Let's look at how sample size concentrates the likelihood function around the true parameter.

```{r exp-concentration, fig.width=8, fig.height=5}
# Calculate log-likelihoods over the grid
ll_small <- sapply(theta_grid_exp, log_lik_exp, data=data_small)
ll_large <- sapply(theta_grid_exp, log_lik_exp, data=data_large)

# Scale them so we can plot them on the same axis easily for shape comparison
# We subtract the max to shift the peak to 0
ll_small_scaled <- ll_small - max(ll_small)
ll_large_scaled <- ll_large - max(ll_large)

plot(theta_grid_exp, ll_small_scaled, type="l", col="blue", lwd=2,
     main="Concentration of Log-Likelihood by Sample Size",
     xlab=expression(theta), ylab="Scaled Log-Likelihood")
lines(theta_grid_exp, ll_large_scaled, col="red", lwd=2)
abline(v = true_theta, lty=2, lwd=2)
legend("bottomright", legend=c("n = 10", "n = 1000", "True Theta"),
       col=c("blue", "red", "black"), lty=c(1, 1, 2), lwd=2)
```

Notice how much more heavily concentrated (sharper peak) the red curve ($n = 1000$) is around the true value compared to the blue curve ($n = 10$).

**Comparing Estimates to Truth:**
The MLE for the exponential parameter is $\widehat{\theta} = \frac{1}{\bar{X}}$.

```{r exp-estimates}
est_small <- 1 / mean(data_small)
est_large <- 1 / mean(data_large)

cat("True Theta:", true_theta, "\n")
cat("Estimate (n=10):", est_small, "\n")
cat("Estimate (n=1000):", est_large, "\n")
```

### Part C: Sample-to-Sample Variation in the Maximum

A core property of the MLE is consistency: as sample size grows, the estimator converges to the true parameter value, meaning the maximum of the likelihood function varies very little across different random samples from the same population.

To see this behavior, we simulate multiple independent datasets (replications) under both a small sample size ($n = 10$) and a large sample size ($n = 1000$), keeping the true rate parameter fixed at $\theta = 2$.

```{r sim-replicates-setup}
set.seed(135)
n_reps <- 5
true_theta <- 2
theta_grid_exp <- seq(0.5, 4.5, by = 0.02)
```

**Small Sample Size ($n = 10$):**
Notice how the individual likelihood curves shift significantly along the x-axis, and the resulting maximums ($\widehat{\theta} = 1 / \bar{X}$) bounce around widely from one realization to the next.

```{r sim-replicates-small, fig.width=7, fig.height=4}
plot(theta_grid_exp, rep(0, length(theta_grid_exp)), type = "n", 
     ylim = c(-15, 0), xlab = expression(theta), ylab = "Scaled Log-Likelihood",
     main = "n = 10: Substantial Variation Across Samples")

palette_cols <- c("firebrick", "dodgerblue", "darkgreen", "darkorange", "purple4")
for (i in 1:n_reps) {
  sample_data <- rexp(10, rate = true_theta)
  ll <- sapply(theta_grid_exp, log_lik_exp, data = sample_data)
  lines(theta_grid_exp, ll - max(ll), col = palette_cols[i], lwd = 2)
  abline(v = 1 / mean(sample_data), col = palette_cols[i], lty = 3, lwd = 4)
}
abline(v = true_theta, lwd = 2, lty = 2)
legend("bottomright", legend = c("Replicate Likelihoods", "True Theta"),
       col = c("purple", "black"), lty = c(1, 2), lwd = c(2, 2))
```

**Large Sample Size ($n = 1000$):**
With a larger sample size, two things happen simultaneously: each likelihood curve becomes noticeably sharper around its peak, and the sample-to-sample estimates tightly cluster directly around the true parameter line ($\theta = 2$).

```{r sim-replicates-large, fig.width=7, fig.height=4}
plot(theta_grid_exp, rep(0, length(theta_grid_exp)), type = "n", 
     ylim = c(-15, 0), xlab = expression(theta), ylab = "Scaled Log-Likelihood",
     main = "n = 1000: Minimal Variation Across Samples")

palette_cols <- c("firebrick", "dodgerblue", "darkgreen", "darkorange", "purple4")
for (i in 1:n_reps) {
  sample_data <- rexp(1000, rate = true_theta)
  ll <- sapply(theta_grid_exp, log_lik_exp, data = sample_data)
  lines(theta_grid_exp, ll - max(ll), col = palette_cols[i], lwd = 2)
  abline(v = 1 / mean(sample_data), col = palette_cols[i], lty = 3, lwd = 4)
}
abline(v = true_theta, lwd = 2, lty = 2)
legend("bottomright", legend = c("Replicate Likelihoods", "True Theta"),
       col = c("purple", "black"), lty = c(1, 2), lwd = c(2, 2))
```

---

### Part D: Optimizing the Likelihood (Calculus and Curvature)

To find the MLE, we use calculus to maximize the log-likelihood function. Let us visualize the geometry of this optimization for an observed sample of exponential data.

```{r exp-calculus-curvature, fig.width=8, fig.height=5}
# Generate a representative sample
set.seed(42)
sample_exp <- rexp(25, rate = 2.0)
mle_exp <- 1 / mean(sample_exp)

# Grid of theta values
theta_eval <- seq(0.5, 4.5, by = 0.01)
log_lik_curve <- sapply(theta_eval, log_lik_exp, data = sample_exp)

# Plot log-likelihood curve
plot(theta_eval, log_lik_curve, type = "l", lwd = 2.5, col = "navy",
     main = "Geometry of the Maximum: Zero Slope and Negative Curvature",
     xlab = expression(theta), ylab = expression(l(theta)))

# Tangent line at the maximum (slope = 0)
max_val <- log_lik_exp(mle_exp, sample_exp)
points(mle_exp, max_val, pch = 19, col = "firebrick", cex = 1.4)
abline(h = max_val, lty = 2, col = "firebrick", lwd = 1.5)
abline(v = mle_exp, lty = 3, col = "gray40", lwd = 4)

# Annotations
text(mle_exp + 0.6, max_val - 3, 
     labels = bquote(widehat(theta)[MLE] == .(round(mle_exp, 3))), 
     col = "black")
text(mle_exp + 0.8, max_val + 1, 
     labels = "Horizontal Tangent: l'("~theta~" | Data) = 0", 
     col = "firebrick", cex = 0.9)
```

**Discussion Questions:**

* **The First-Derivative Test:**
  * Looking at the tangent line above, what is the geometric meaning of setting $\frac{d}{d\theta} \ell(\theta) = 0$?
  * Does finding a point where the first derivative equals zero guarantee that we have found a maximum, or could it represent something else?

* **Curvature and Second Derivatives:**
  * For the Exponential log-likelihood $\ell(\theta) = n\log(\theta) - \theta \sum_{i=1}^n X_i$, the second derivative is:
    $$\frac{d^2}{d\theta^2} \ell(\theta) = -\frac{n}{\theta^2}$$
  * What is the sign of this second derivative for all valid $\theta > 0$? 
  * Why does a strictly negative second derivative assure us that the peak is a *global* maximum, eliminating concerns about local optima?

* **When Calculus Breaks Down:**
  * This calculus optimization assumes that the log-likelihood is smooth and differentiable everywhere. What would happen if our likelihood had a sharp corner (like an absolute value function / $L_1$ loss), or if the support of the distribution depended on the parameter itself (like $U(0, \theta)$)?
  * Does the absence of a zero derivative mean no maximum exists, or does it mean we need a different maximization strategy?
  
### When Calculus Fails: Non-Differentiability and $L_1$ Loss

Setting the derivative equal to zero only works when the likelihood function is differentiable everywhere. When a function has sharp corners or boundaries, an optimum can still exist even if no zero derivative can be found via standard calculus.

```{r non-diff-plot, fig.width=6, fig.height=3.5}
# Illustrating non-differentiability with the absolute value function |theta|
theta_vals <- seq(-3, 3, length.out = 300)
loss_vals <- abs(theta_vals)

plot(theta_vals, loss_vals, type = "l", lwd = 2.5, col = "darkred",
     main = expression("Non-Differentiable Minimum: " * f(theta) == abs(theta)),
     xlab = expression(theta), ylab = expression(abs(theta)))
points(0, 0, pch = 19, col = "black", cex = 1.3)
abline(v = 0, lty = 3, col = "gray50")
text(0, 0.5, "Sharp minimum at 0\n(Derivative undefined)", pos = 4, cex = 0.85)
```

* **Takeaway:** A function does not need to be differentiable to have an optimum. At $\theta = 0$, the absolute value function clearly achieves its minimum, but $\frac{d}{d\theta}\vert{}\theta\vert{}$ does not exist at that point.
* **Connection to $L_1$ Loss:** This behavior occurs in $L_1$ loss optimization ($\sum \vert{}x_i - \theta\vert{}$), as well as distributions with bounded support like $U(0, \theta)$ where the boundary depends on the parameter.

### Estimator vs. Estimate

In statistical inference, we must formally distinguish between an **estimator** (a statistic defined as a function of the random variables) and an **estimate** (the fixed numerical value obtained by evaluating that function on an observed sample):

* **Estimator (The Statistic):** A function of the random sample $X_1, \dots, X_n$:
  $$\widehat{\theta}(X_1, \dots, X_n) = \frac{1}{\bar{X}} = \frac{n}{\sum_{i=1}^n X_i}$$
  Because it depends on random variables, $\widehat{\theta}$ is itself a random variable endowed with a sampling distribution, mean, and variance. The data arguments are written with capital letters ($X_i, \bar{X}$) to reflect that the estimator is a function of unobserved random variables before the experiment is run.

* **Estimate (The Realization):** The fixed numerical value obtained when the estimator is evaluated on a concrete, observed data vector $(x_1, \dots, x_n)$:
  $$\widehat{\theta}_{\text{obs}} = \frac{1}{\bar{x}} = \frac{n}{\sum_{i=1}^n x_i}$$
  An estimate is a fixed constant with no sampling variability, written with lowercase letters ($x_i, \bar{x}$) to denote specific, realized numbers recorded from the experiment.

```{r compare-estimates}
# True population rate
true_theta <- 2.0

# Generate one small sample (n = 10) and one large sample (n = 1000)
set.seed(42)
sample_10  <- rexp(10, rate = true_theta)
sample_1000 <- rexp(1000, rate = true_theta)

# Plug observed sample data into our estimator rule to obtain concrete estimates
est_10  <- 1 / mean(sample_10)
est_1000 <- 1 / mean(sample_1000)

cat("True parameter (theta):", true_theta, "\n")
cat("Estimate with n = 10:   ", round(est_10, 3), "\n")
cat("Estimate with n = 1000:  ", round(est_1000, 3), "\n")
```

---

## 3. Attendance Question: Evaluating Model Assumptions

**Question:**
In the exponential example, what problem would you have with your MLE estimator if you had the following data?
$X_1 = 1, X_2 = 0, X_3 = 3, X_4 = 2, X_5 = -6$?

**Analysis:**
If we naively apply our plug-in formula $\widehat{\theta} = \frac{1}{\bar{X}}$, we immediately encounter an undefined result because the sample mean sums to zero ($\bar{X} = \frac{1 + 0 + 3 + 2 - 6}{5} = 0$). 

However, the fundamental breakdown occurs at the modeling stage: **an exponential distribution is only defined on strictly positive values, so it cannot generate negative data**.

**The Indicator Function:**
To capture the valid domain of a distribution mathematically, we use an **indicator function** $I(\cdot)$ directly in the density:
$$f(x) = \theta \exp(-\theta x) I(x > 0), \quad \text{where } I(x > 0) = \begin{cases} 1 & \text{if } x > 0 \\ 0 & \text{if } x \le 0 \end{cases}$$

When we write out the joint likelihood for the full sample, the individual indicators multiply together:
$$\ell(\theta) = \prod_{i=1}^n \Big(\theta \exp(-\theta X_i) I(X_i > 0)\Big) = I(\text{all } X_i > 0) \prod_{i=1}^n \theta \exp(-\theta X_i)$$

* Because $X_5 = -6 \le 0$, the condition "$\text{all } X_i > 0$" fails, so $I(\text{all } X_i > 0) = 0$.
* Consequently, the likelihood is **identically 0 for every candidate value of $\theta$**, meaning no maximum exists.
* This makes sense: observing negative values is impossible under an exponential model, yielding zero likelihood.

### Why Indicators Matter: The $\text{Uniform}(0, \theta)$ Example

Including indicators is not just formal bookkeeping; it is essential whenever a parameter restricts the support (boundaries) of the distribution. In these cases, standard calculus fails, and the indicator function is what determines the MLE.

Consider a sample $X_1, \dots, X_n \overset{\text{i.i.d.}}{\sim} \text{Uniform}(0, \theta)$:

**The Density Requires an Indicator:**
The uniform density is zero outside the interval $[0, \theta]$, so we write:
$$f(x) = \frac{1}{\theta} I(0 \le x \le \theta) \quad \text{for } \theta > 0$$

**The Joint Likelihood Enforces a Constraint on the Parameter:**
Multiplying across all $n$ independent observations gives:
$$L(\theta) = \prod_{i=1}^n \left( \frac{1}{\theta} I(0 \le X_i \le \theta) \right) = \frac{1}{\theta^n} I(\text{all } 0 \le X_i \le \theta) = \frac{1}{\theta^n} I(\max(X_1, \dots, X_n) \le \theta)$$
*(Every individual observation is $\le \theta$ if and only if the sample maximum is $\le \theta$.)*

**Why Standard Calculus Fails:**
Taking the derivative of $\frac{1}{\theta^n}$ with respect to $\theta$ yields:
$$\frac{d}{d\theta}\left(\frac{1}{\theta^n}\right) = -\frac{n}{\theta^{n+1}}$$
Because this derivative is strictly negative for all $\theta > 0$, it is never zero. Calculus alone cannot find a critical point.

**Finding the Maximum via Boundary Logic:**

* The term $\frac{1}{\theta^n}$ is strictly decreasing for $\theta > 0$, so we maximize the function by choosing $\theta$ to be as **small** as possible.
* However, the indicator function imposes a sharp lower boundary: if we choose $\theta < \max(X_1, \dots, X_n)$, the indicator evaluates to $0$, causing the entire likelihood to drop to zero.
* Therefore, the smallest value $\theta$ can take without zeroing out the likelihood is the largest observed data point itself:
  $$\widehat{\theta} = \max(X_1, \dots, X_n)$$

**Key Takeaway:**
Always check the support of your observed data against the assumptions of your distribution. A parametric model must be capable of generating the data you actually observed.
