---
title: "Maximum Likelihood with Multiple Parameters"
subtitle: "Two-Group Normal Models and Gamma Optimization"
author: "STAT 135 Lab 3"
date: "Fall 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5, fig.align = "center"
)
set.seed(13503)
```

In Lab 2, we maximized likelihoods with one unknown parameter. Now we
will estimate several parameters at once. Sometimes a problem separates
into smaller problems with explicit solutions. In other cases, we need
numerical optimization.

Run the chunks in order. The statistical calculations and plots use
base R; `knitr` formats the tables. Throughout this lab, `log` means the
natural logarithm. We use $\ell(\theta)$ for likelihood and $L(\theta)$ for log-likelihood, so $L(\theta)=\log\ell(\theta)$.

## 1. A Two-Group Normal Model

### Step 1: Specify the model and its assumptions

Suppose $Y$ is a continuous response and $X$ identifies one of two groups:
$X=0$ or $X=1$. For example, $Y$ might be a measurement collected under
each of two experimental conditions.

We model the response conditional on group membership as
$$
Y_i\mid X_i=x_i \sim
\begin{cases}
\mathcal{N}(\alpha_0,\sigma_0^2), & x_i=0,\\
\mathcal{N}(\alpha_1,\sigma_1^2), & x_i=1.
\end{cases}
$$
We assume the responses are independent conditional on the observed group
labels. Within each group, responses share the same normal distribution.
We allow both the mean and the variance to differ between groups.

There are **four unknown parameters**: $\alpha_0$, $\alpha_1$,
$\sigma_0^2>0$, and $\sigma_1^2>0$. We condition on the observed $X$
values, so we do not need a probability model for group membership.

**Question:** Does having two different group means necessarily imply
that the groups have different variances? What does each parameter describe?

For illustration, simulate two groups with different means and spreads.
These are simulated measurements, not an actual experiment.

```{r two-group-data}
set.seed(13503)
n0 <- 30
n1 <- 40
x <- c(rep(0, n0), rep(1, n1))
y <- c(rnorm(n0, mean = 10, sd = 2),
       rnorm(n1, mean = 14, sd = 3))

y0 <- y[x == 0]
y1 <- y[x == 1]

boxplot(y ~ x, col = c("lightblue", "mistyrose"),
        names = c("Group 0", "Group 1"),
        main = "Two Groups: Location and Spread",
        xlab = "Group", ylab = "Response Y")
points(c(1, 2), c(mean(y0), mean(y1)), pch = 19,
       col = "firebrick", cex = 1.3)
legend("topleft", legend = "Sample mean", pch = 19,
       col = "firebrick", bty = "n")
```

### Step 2: Write the likelihood and log-likelihood

For simplicity, label the observations so that the first $n_0$ belong
to group 0 and the next $n_1$ belong to group 1. Define
$$
I_0=\{1,\ldots,n_0\},\qquad
I_1=\{n_0+1,\ldots,n_0+n_1\}.
$$
This is just a relabeling of the data: each response stays paired with
its group label. It does not change the likelihood, the estimates, or
any of the calculations below; it only simplifies the notation.

Independence lets us multiply the individual normal densities:
$$
\ell(\alpha_0,\alpha_1,\sigma_0^2,\sigma_1^2)
=\prod_{g=0}^1\prod_{i\in I_g}
\frac{1}{\sqrt{2\pi\sigma_g^2}}
\exp\left\{-\frac{(y_i-\alpha_g)^2}{2\sigma_g^2}\right\}.
$$
Taking logs turns the product into a sum:
$$
L=L_0(\alpha_0,\sigma_0^2)+L_1(\alpha_1,\sigma_1^2),
$$
where
$$
L_g(\alpha_g,\sigma_g^2)
=-\frac{n_g}{2}\log(2\pi)
-\frac{n_g}{2}\log(\sigma_g^2)
-\frac{1}{2\sigma_g^2}\sum_{i\in I_g}(y_i-\alpha_g)^2.
$$

**Key observation:** The group 0 contribution does not involve the group 1
parameters, and vice versa. Maximizing the full log-likelihood therefore
reduces to solving two separate normal MLE problems.

**Question:** When differentiating $L$ with respect to $\alpha_0$,
what happens to the entire group 1 contribution?

### Step 3: Use partial derivatives

A partial derivative varies one parameter while holding the other
parameters fixed. For either group $g$, differentiating with respect to
its mean gives
$$
\frac{\partial L_g}{\partial\alpha_g}
=\frac{1}{\sigma_g^2}\sum_{i\in I_g}(y_i-\alpha_g).
$$
Setting this equal to zero yields
$$
\widehat\alpha_g=\bar y_g
=\frac{1}{n_g}\sum_{i\in I_g}y_i.
$$
The second derivative is $-n_g/\sigma_g^2<0$, so for any fixed positive
variance, the sample mean maximizes the log-likelihood over $\alpha_g$.

To differentiate with respect to the **variance**, write $v_g=\sigma_g^2$:
$$
\frac{\partial L_g}{\partial v_g}
=-\frac{n_g}{2v_g}
+\frac{\sum_{i\in I_g}(y_i-\alpha_g)^2}{2v_g^2}.
$$
Substituting $\widehat\alpha_g=\bar y_g$ and solving gives
$$
\widehat\sigma_g^2
=\frac{1}{n_g}\sum_{i\in I_g}(y_i-\bar y_g)^2.
$$
If the within-group sum of squared deviations is positive, the derivative
in $v_g$ is positive below this value and negative above it. Thus it is
the maximizing variance. We assume each group has at least two
observations with some within-group variation; with all values identical,
the likelihood grows without bound as the variance approaches zero.

**Exercise:** Write out the four estimates explicitly for groups 0 and 1.
Which observations contribute to each estimate?

### Step 4: Calculate the estimates in R

```{r normal-mles}
alpha0_hat <- mean(y0)
alpha1_hat <- mean(y1)
sigma20_hat <- mean((y0 - alpha0_hat)^2)
sigma21_hat <- mean((y1 - alpha1_hat)^2)

normal_results <- data.frame(
  group = c(0, 1), n = c(length(y0), length(y1)),
  true_mean = c(10, 14),
  estimated_mean = c(alpha0_hat, alpha1_hat),
  true_variance = c(4, 9),
  estimated_variance = c(sigma20_hat, sigma21_hat)
)
knitr::kable(normal_results, digits = 3)
```

The MLE variance uses divisor $n_g$, whereas R's `var()` uses $n_g-1$.
This is the same distinction we encountered in Lab 1. Under the model,
`var()` is unbiased for the group variance; the MLE variance is biased
downward in finite samples.

```{r variance-divisors}
knitr::kable(data.frame(
  group = c(0, 1),
  MLE_variance = c(sigma20_hat, sigma21_hat),
  R_var = c(var(y0), var(y1)),
  converted_R_var = c((n0 - 1) / n0 * var(y0),
                      (n1 - 1) / n1 * var(y1))
), digits = 3)
```

**Question:** If we instead assumed $\sigma_0^2=\sigma_1^2$, could we
still estimate two unrelated variances? Why would both groups now
contribute to estimating the shared variance?

## 2. The Gamma Model: Two Parameters That Must Be Solved Together

### Step 1: Review the distribution

The Gamma distribution can model positive, continuous measurements such
as waiting times. Here $\alpha>0$ is the **shape** and $\lambda>0$ is the
**rate**. These symbols refer to Gamma parameters, not the group means
from Section 1.

The density for $z>0$ is
$$
f(z;\alpha,\lambda)
=\frac{\lambda^\alpha}{\Gamma(\alpha)}
z^{\alpha-1}e^{-\lambda z},
\qquad
\mathrm{E}(Z)=\frac{\alpha}{\lambda},\quad
\mathrm{Var}(Z)=\frac{\alpha}{\lambda^2}.
$$
The density is zero for $z<0$; the value assigned at the single point
zero does not affect continuous probabilities. Our likelihood
calculations below use strictly positive observations because they
involve $\log z_i$.

The **Gamma function** is
$$
\Gamma(\alpha)=\int_0^\infty t^{\alpha-1}e^{-t}\,dt.
$$
It extends the factorial: for positive integers $k$,
$\Gamma(k)=(k-1)!$. It also provides the normalizing constant that
makes the Gamma density integrate to one.

```{r gamma-function}
c(Gamma_1 = gamma(1), Gamma_5 = gamma(5), factorial_4 = factorial(4))
```

R uses `shape` and `rate` in `dgamma()` and `rgamma()`. Some references
use a scale parameter instead; **scale is the reciprocal of rate**.

```{r gamma-densities}
z_grid <- seq(0.001, 10, length.out = 500)
shapes <- c(1, 2, 5)
cols <- c("firebrick", "dodgerblue", "darkgreen")
plot(z_grid, dgamma(z_grid, shape = shapes[1], rate = 1),
     type = "l", lwd = 2, col = cols[1], ylim = c(0, 1),
     main = "Gamma Densities with Rate = 1",
     xlab = "z", ylab = "Density")
for (j in 2:length(shapes)) {
  lines(z_grid, dgamma(z_grid, shape = shapes[j], rate = 1),
        col = cols[j], lwd = 2)
}
legend("topright", legend = paste("Shape =", shapes),
       col = cols, lwd = 2, bty = "n")
```

When shape equals 1, Gamma is the exponential model from Lab 2.
Increasing shape at a fixed rate changes both the mean and variance.

**Discussion question:** What characteristics would you want in your
data to make Gamma a reasonable choice?

We would look for the following characteristics:

* **Positive values:** Gamma models measurements greater than zero,
  such as waiting times or lifetimes. Negative observations would be
  incompatible with the model. If recorded zeros occur, we should ask
  whether they reflect rounding or a separate process before using the
  strictly positive-data likelihood below.
* **Continuous measurements:** Values should vary along a continuous
  scale, such as time in minutes, rather than represent counts or binary
  outcomes. Rounding a time measurement does not make the underlying
  quantity a count.
* **A plausible distributional shape:** Gamma distributions are
  right-skewed, with a longer tail toward large values. For shape at most
  1, the density decreases from the boundary at zero; for shape greater
  than 1, it has one interior peak. Large shape parameters produce less
  skewness, so the data need not look strongly skewed. Several distinct
  peaks would suggest that a single Gamma distribution may be unsuitable.
* **Reasonable independence and common parameters:** For our iid model,
  observations should not strongly depend on one another and should
  plausibly come from the same population with a common shape and rate.
  Repeated measurements on the same subject or a mixture of groups with
  different typical values may require a different model.

For example, positive waiting times from comparable, independent events,
with many moderate waits and a few long waits, could be reasonable
candidates. These characteristics motivate a Gamma model but do not
prove it fits; compare the observed distribution with the fitted density.

### Step 2: Derive the log-likelihood

Suppose $Z_1,\ldots,Z_n$ are iid Gamma$(\alpha,\lambda)$. For observed
values $z_1,\ldots,z_n>0$,
$$
\ell(\alpha,\lambda)
=\frac{\lambda^{n\alpha}}{\Gamma(\alpha)^n}
\left(\prod_{i=1}^n z_i^{\alpha-1}\right)
\exp\left(-\lambda\sum_{i=1}^n z_i\right).
$$
Therefore,
$$
L(\alpha,\lambda)
=n\alpha\log\lambda-n\log\Gamma(\alpha)
+(\alpha-1)\sum_{i=1}^n\log z_i
-\lambda\sum_{i=1}^n z_i.
$$
Unlike the two-group example, this expression does not separate into
one function of $\alpha$ plus another function of $\lambda$. The term
$n\alpha\log\lambda$ involves both parameters.

**Exercise:** Derive this expression from the product of individual
densities. Which terms contain both unknown parameters?

### Step 3: Find the likelihood equations

Differentiating with respect to rate gives
$$
\frac{\partial L}{\partial\lambda}
=\frac{n\alpha}{\lambda}-\sum_{i=1}^n z_i.
$$
Setting it equal to zero gives a relationship between the estimates:
$$
\widehat\lambda=\frac{\widehat\alpha}{\bar z}.
$$
This is not yet a numerical answer: $\widehat\alpha$ is still unknown.

Differentiating with respect to shape gives
$$
\frac{\partial L}{\partial\alpha}
=n\log\lambda-n\frac{\Gamma'(\alpha)}{\Gamma(\alpha)}
+\sum_{i=1}^n\log z_i.
$$
The function $\psi(\alpha)=\Gamma'(\alpha)/\Gamma(\alpha)$ is called the
**digamma function**; R evaluates it using `digamma()`.
Substituting the rate equation and dividing by $n$ yields
$$
\log\widehat\alpha-\psi(\widehat\alpha)
=\log\bar z-\frac{1}{n}\sum_{i=1}^n\log z_i.
$$
There is no general closed-form expression for $\widehat\alpha$.
We can still estimate both parameters by numerical optimization.

**Question:** Why does finding a formula for $\widehat\lambda$ in terms
of $\widehat\alpha$ not finish the problem?

### Step 4: Simulate data and choose starting values

```{r gamma-data}
set.seed(13504)
n_gamma <- 150
true_shape <- 3
true_rate <- 1.5
z <- rgamma(n_gamma, shape = true_shape, rate = true_rate)

c(sample_mean = mean(z), sample_variance = var(z),
  true_mean = true_shape / true_rate,
  true_variance = true_shape / true_rate^2)
```

An iterative optimizer needs initial guesses. From the mean and variance
formulas, $\alpha=\mu^2/v$ and $\lambda=\mu/v$. Substituting the sample
mean and the empirical variance gives useful starting values.
These moment-based guesses are not generally the MLEs.

```{r gamma-start}
m <- mean(z)
v <- mean((z - m)^2)
start <- c(shape = m^2 / v, rate = m / v)
start
```

### Step 5: Use `optim()` to maximize the log-likelihood

By default, `optim()` **minimizes** a function, so we supply the
**negative log-likelihood**. Adding log densities is more numerically
stable than multiplying many small density values.

Both parameters must be positive. We optimize their logarithms:
$\eta_1=\log\alpha$ and $\eta_2=\log\lambda$. Exponentiating inside the
function ensures that every proposed shape and rate is positive.
This just changes how we describe the parameters; the data likelihood
remains the same.

```{r gamma-optim}
neg_loglik <- function(log_par, data) {
  shape <- exp(log_par[1])
  rate <- exp(log_par[2])
  -sum(dgamma(data, shape = shape, rate = rate, log = TRUE))
}

fit_gamma <- optim(
  par = log(start),
  fn = neg_loglik,
  data = z,
  method = "BFGS"
)

gamma_hat <- exp(fit_gamma$par)
alpha_hat <- unname(gamma_hat["shape"])
lambda_hat <- unname(gamma_hat["rate"])

knitr::kable(data.frame(
  parameter = c("Shape", "Rate"),
  truth = c(true_shape, true_rate),
  starting_value = as.numeric(start),
  MLE = as.numeric(gamma_hat)
), digits = 4)

c(convergence_code = fit_gamma$convergence,
  negative_loglik_at_solution = fit_gamma$value)
```

`par` gives the starting log parameters; `fn` is the function to minimize;
`data` passes our observations to that function. `BFGS` is an iterative
optimization method. The returned `par` contains the fitted **log**
parameters, so we use `exp()` to recover shape and rate.

A convergence code of 0 indicates successful completion according to
the optimizer's criteria. It is useful to check the likelihood equations
as well, rather than relying only on that code.

```{r gamma-checks}
c(
  fitted_rate = lambda_hat,
  shape_over_sample_mean = alpha_hat / mean(z),
  shape_equation_residual = log(alpha_hat) - digamma(alpha_hat) -
    (log(mean(z)) - mean(log(z))),
  loglik_improvement = neg_loglik(log(start), z) - fit_gamma$value
)
```

These quantities check whether the numerical estimates satisfy the
likelihood equations we derived:

* **`shape_over_sample_mean`** computes $\widehat\alpha/\bar z$.
  Setting the derivative with respect to rate equal to zero gave
  $\widehat\lambda=\widehat\alpha/\bar z$. Therefore, this value should
  agree closely with `fitted_rate`, the rate estimate returned by
  `optim()`.
* **`shape_equation_residual`** computes the left side minus the right
  side of our shape equation:
  $$
  \underbrace{\log\widehat\alpha-\psi(\widehat\alpha)}_{\text{left side}}
  -\underbrace{\left(\log\bar z-
  \frac{1}{n}\sum_{i=1}^n\log z_i\right)}_{\text{right side}}.
  $$
  Here $\psi$ is the digamma function, evaluated by R's `digamma()`.
  At the exact solution, the two sides are equal, so the residual is
  zero. A residual close to zero indicates that the fitted shape
  approximately satisfies this equation.
* **`loglik_improvement`** is the fitted log-likelihood minus the
  starting log-likelihood. It should be nonnegative up to numerical
  rounding, since optimization should improve on the starting guess.

The first two checks together assess whether `optim()` approximately
solved both likelihood equations. Small discrepancies are expected
because numerical optimization uses finite precision and stopping
tolerances. These are checks of the optimization, not proof that the
Gamma model is a good fit to the data.

### Step 6: Visualize the fitted model and likelihood

```{r gamma-fitted-density}
hist(z, probability = TRUE, breaks = "FD", col = "gray85",
     border = "white", main = "Gamma Data and Fitted Density",
     xlab = "Observed value")
curve(dgamma(x, shape = alpha_hat, rate = lambda_hat),
      add = TRUE, col = "firebrick", lwd = 2)
curve(dgamma(x, shape = true_shape, rate = true_rate),
      add = TRUE, col = "navy", lwd = 2, lty = 2)
legend("topright", legend = c("Fitted density", "True density"),
       col = c("firebrick", "navy"), lwd = 2, lty = c(1, 2),
       bty = "n")
```

The true density is available here because we simulated the data.
With real observations, we would not know the true parameters.
The histogram is on a density scale so that the overlays are comparable.

With two parameters, a contour plot lets us view the log-likelihood
as a surface. Each contour connects parameter pairs with the same
log-likelihood value.

```{r gamma-likelihood-contours}
shape_grid <- seq(0.6 * alpha_hat, 1.4 * alpha_hat, length.out = 80)
rate_grid <- seq(0.6 * lambda_hat, 1.4 * lambda_hat, length.out = 80)
ll_grid <- outer(shape_grid, rate_grid, Vectorize(function(a, r) {
  sum(dgamma(z, shape = a, rate = r, log = TRUE))
}))

# Shift by the fitted maximum; this does not change the maximizing pair.
relative_ll <- ll_grid + fit_gamma$value
contour(shape_grid, rate_grid, relative_ll,
        levels = c(-20, -10, -5, -2, -0.5),
        xlab = expression(alpha~"(shape)"),
        ylab = expression(lambda~"(rate)"),
        main = "Gamma Log-Likelihood Relative to Its Maximum")
points(alpha_hat, lambda_hat, col = "firebrick", pch = 19)
points(true_shape, true_rate, col = "navy", pch = 4, lwd = 2)
legend("topleft", legend = c("MLE", "Truth"),
       col = c("firebrick", "navy"), pch = c(19, 4), bty = "n")
```

**Discussion:** Why is the fitted maximum not exactly at the true
parameter pair? If we generated another sample, would the contour plot
and estimates remain identical?

**Try it:** Rerun `optim()` starting from `log(c(shape = 1, rate = 1))`.
Compare the estimates, convergence code, and final objective value.
Agreement from different starts provides a useful numerical check.

## 3. Attendance Question: A Binary Predictor

Return to the two-group setting from Section 1. Attempt the attendance
question before going through the solution below.

Our standard linear model is
$$
\mathrm{E}(Y\mid X=x)=\beta_0+\beta_1x.
$$
Suppose the predictor $X$ is binary, either 0 or 1.

1. What is $\mathrm{E}(Y\mid X=0)$ in this model?
2. What is $\mathrm{E}(Y\mid X=1)$ in this model?

### End-of-lab walkthrough

Substitute the two possible predictor values:
$$
\mathrm{E}(Y\mid X=0)=\beta_0+\beta_1(0)=\beta_0,
$$
$$
\mathrm{E}(Y\mid X=1)=\beta_0+\beta_1(1)=\beta_0+\beta_1.
$$
Thus we can rewrite the conditional mean as
$$
\mathrm{E}(Y\mid X=x)=
\begin{cases}
\beta_0, & x=0,\\
\beta_0+\beta_1, & x=1.
\end{cases}
$$
Comparing with our earlier group-mean notation gives
$$
\alpha_0=\beta_0,\qquad \alpha_1=\beta_0+\beta_1,
$$
or equivalently,
$$
\beta_0=\alpha_0,\qquad \beta_1=\alpha_1-\alpha_0.
$$

These are two ways to describe the same conditional means, but the
coefficients have different meanings:

* $\alpha_0$ and $\beta_0$ are the mean response in group 0.
* $\alpha_1$ is the mean response in group 1.
* $\beta_1$ is the difference in mean response: group 1 minus group 0.

```{r attendance-connection}
beta0_hat <- alpha0_hat
beta1_hat <- alpha1_hat - alpha0_hat

knitr::kable(data.frame(
  group = c(0, 1),
  fitted_group_mean = c(alpha0_hat, alpha1_hat),
  fitted_linear_mean = beta0_hat + beta1_hat * c(0, 1)
), digits = 3)

# With a binary predictor, least squares gives the same fitted means.
fit_linear <- lm(y ~ x)
c(intercept = unname(coef(fit_linear)[1]),
  slope = unname(coef(fit_linear)[2]))
```

This equality concerns the fitted **means**. Our original normal model
allows different group variances; the usual constant-variance normal
regression model imposes an additional assumption. Matching the mean
formulas does not require the variances to be equal.

**Final check:** If the group means were 10 and 14, what would
$(\alpha_0,\alpha_1)$ be? What would $(\beta_0,\beta_1)$ be? Explain why
$\beta_1$ is a difference rather than the mean of group 1.
