---
title: "Sampling Distributions and Estimation"
subtitle: "The Shifted Exponential Model"
author: "STAT 135 Lab 4"
date: "Fall 2026"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)

set.seed(13504)
```

# The Shifted Exponential Model

Let $Y\sim\operatorname{Exp}(\lambda)$, where rate $\lambda>0$, and define $X=\theta+Y$. The parameter $\theta\in\mathbb R$ shifts the support to start at $\theta$ without changing its shape:

$$
f(x;\theta,\lambda)=
\begin{cases}
\lambda e^{-\lambda(x-\theta)},&x\ge\theta,\\
0,&x<\theta.
\end{cases}
$$

Adding a constant shifts the mean but does not change the variance:

$$
\mathrm E(X)=\theta + \mathrm E(Y) = \theta+\frac1\lambda,\qquad
\operatorname{Var}(X)=\operatorname{Var}(Y) = \frac1{\lambda^2}.
$$

```{r shifted-densities, fig.width=9, fig.height=4}
lambda <- 0.1
shifts <- c(0, 5, 15)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1), oma = c(0, 0, 3, 0))

for (shift in shifts) {

  plot(
    NA,
    xlim = c(-5, 60),
    ylim = c(0, 0.12),
    xlab = "x",
    ylab = "Density",
    main = paste("Shift =", shift)
  )

  # Density is 0 for x < shift
  segments(
    x0 = -5, y0 = 0,
    x1 = shift, y1 = 0,
    lwd = 2
  )

  # Density for x >= shift
  x <- seq(shift, 60, length.out = 500)

  density <- lambda * exp(
    -lambda * (x - shift)
  )

  lines(x, density, lwd = 2)

  # Show the jump at x = shift
  points(shift, 0, pch = 21, bg = "white")
  points(shift, lambda, pch = 19)

  abline(v = shift, lty = 2, col = "gray")
}

mtext(
  expression(lambda == 0.1 * ", Different Shifts"),
  side = 3,
  outer = TRUE,
  line = 1,
  font = 2
)
```

**Question:** With $\theta=5$ and $\lambda=0.1$, what are the mean, variance, and standard deviation of one observation? Which change if we increase only $\theta$?

# Maximum Likelihood with A Support Constraint

Suppose

$$
X_1,\ldots,X_n
\stackrel{\text{iid}}{\sim}
\text{shifted exponential}(\theta,\lambda).
$$

For observed data $x_1,\ldots,x_n$, the likelihood is

$$
\ell(\theta,\lambda)
=\lambda^n\exp\left\{-\lambda\sum_{i=1}^n(x_i-\theta)\right\}
I\{x_i \ge \theta\},\qquad \lambda>0.
$$The condition $x_i\ge\theta$ must hold for *every* observation. Let

$$
x_{(1)}=\min(x_1,\ldots,x_n).
$$

Then all indicators equal one exactly when

$$
\theta\le x_{(1)}.
$$

Therefore,

$$
\ell(\theta,\lambda)
=\lambda^n
\exp\left\{-\lambda\sum_{i=1}^n(x_i-\theta)\right\}
I(\theta\le x_{(1)}).
$$

On the feasible region $\theta\le x_{(1)}$, the log-likelihood is

$$
L(\theta,\lambda)
=n\log\lambda
-\lambda\sum_{i=1}^n x_i
+n\lambda\theta.
$$Outside that region, the likelihood is zero, so the log-likelihood is $-\infty$.

Hence, the log-likelihood is

$$
L(\theta,\lambda)=
\begin{cases}
n\log\lambda-\lambda\sum_{i=1}^n x_i+n\lambda\theta,
  &\theta\le x_{(1)}\\
-\infty,&\theta>x_{(1)}
\end{cases}.
$$

## Maximize over the shift

For any fixed $\lambda>0$, inside the feasible region,

$$
\frac{\partial L}{\partial\theta}=n\lambda>0.
$$There is no interior point where this derivative equals zero. But this does *not* mean the likelihood has no maximum. Since the log-likelihood is strictly increasing as a function of $\theta$ on its constrained domain $\theta\le x_{(1)}$, choose the largest feasible shift:

$$\widehat\theta=x_{(1)}.$$

## Maximize over the rate

Substituting $\widehat\theta=x_{(1)}$ into the log-likelihood and differentiating with respect to $\lambda$ gives

$$
\widehat\lambda
=\frac{n}{\sum_{i=1}^n(x_i-x_{(1)})}
=\frac{1}{\bar x-x_{(1)}}.
$$

The second derivative is $-n/\lambda^2<0$.

Thus,

$$
\boxed{
\widehat\theta=X_{(1)},
\qquad
\widehat\lambda=\frac{1}{\overline X-X_{(1)}}.
}
$$

For the demonstrations below, use

$$
\theta=5,\qquad \lambda=0.1.
$$

```{r parameters}
theta_true <- 5
lambda_true <- 0.1

c(
  theta_true = theta_true,
  lambda_true = lambda_true
)
```

Now, let's draw one sample of size 20 and calculate the corresponding maximum likelihood estimates.

```{r one-sample-mles}
n <- 20
x <- theta_true + rexp(n, rate = lambda_true)

theta_hat <- min(x)
lambda_hat <- 1 / (mean(x) - theta_hat)

knitr::kable(
  data.frame(
    parameter = c("Shift", "Rate"),
    truth = c(theta_true, lambda_true),
    estimate = c(theta_hat, lambda_hat)
    ), 
  digits = 4
)
```

You will find that the estimates $\hat{\theta}$ and $\hat{\lambda}$ change as you repeatedly run the code chunk above.

The **estimator** is a function of the sample data, so it is a random variable. An **estimate** is the numerical value obtained after applying the estimator to a particular observed sample. Therefore, different samples can produce different estimates.

**Question:** Is `theta_hat` possible to be smaller than the true $\theta=5$ under this model?

# Sampling Distribution

Suppose $X_1,\ldots,X_n$ are random variables from a population model with unknown parameter $\theta$. An estimator of $\theta$ is a function of the sample data,

$$
\widehat\theta=T(X_1,\ldots,X_n).
$$

Before the sample data is observed, $\widehat\theta$ is random because it depends on random variables. Its probability distribution across repeated samples of a fixed size is the *sampling distribution* of $\widehat\theta$.

A **sampling distribution** is the distribution of a statistic across repeated samples of a fixed size from the same population. Distinguish:

-   $X_i$: one random variable from the population distribution.

-   $x_i$: the realization (observed value) of $X_i$.

-   $\widehat\theta=T(X_1,\ldots,X_n)$: an estimator (a statistic) — a random variable.

-   $\widehat\theta(x_1,\ldots,x_n)$: an estimate — a numerical value after applying the estimator on the observed sample data.

-   the **data distribution** describes one random variable $X_i$.

-   the **sampling distribution** describes one statistic computed from a whole sample.

## Simulate the sampling distribution of $\hat\theta$

To simulate the sampling distribution of $\hat\theta$ of sample size $20$, we repeatedly draw a sample of size $20$ from the population and calculate the estimate for each sample.

```{r simulate-theta-hat}
B <- 100000
n <- 20

theta_hat_sim <- replicate(B, {
  new_sample <- theta_true + rexp(n, rate = lambda_true)
  min(new_sample)
})

c(
  simulated_mean = mean(theta_hat_sim),
  simulated_variance = var(theta_hat_sim)
)
```

The simulation gives 10,000 estimates of $\theta$, one from each simulated sample.

```{r compare-distributions, fig.height=4}
individual_sim <- theta_true + rexp(B, rate = lambda_true)
plot_limit <- theta_true + qexp(0.975, rate = lambda_true)
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1), cex.main = 0.85)
hist(
  individual_sim, probability = TRUE, breaks = 40,
  xlim = c(theta_true, plot_limit), col = "gray85", border = "white",
  main = expression("Simulated Data Distribution"), xlab = "x"
)
curve(dexp(x - theta_true, rate = lambda_true), add = TRUE,
      col = "firebrick", lwd = 2)

hist(theta_hat_sim, probability = TRUE, breaks = 40,
     xlim = c(theta_true, plot_limit), col = "skyblue", border = "white",
     main = expression("Simulated Sampling Distribution of " * hat(theta)), 
     xlab = expression(hat(theta)))
par(mfrow = c(1, 1))
```

The panels use the same horizontal limits (a few extreme observations may be outside the displayed range). Their vertical scales differ. Both distributions are right-skewed, but the the simulated sampling distribution of $\hat\theta$ is more concentrated.

Can we get the exact sampling distribution of $\hat\theta = X_{(1)}$?

## Exact sampling distribution of $\hat\theta = X_{(1)}$

We want the distribution of the minimum in the sample. Directly calculating

$$
P(X_{(1)}\le u)
$$can be awkward because "the minimum is at most $u$" means at least one observation is at most $u$. The complement is easier:

$$
P(X_{(1)}>u)
=P(X_1>u,\ldots,X_n>u).
$$

For $u\ge\theta$, independence gives

$$
P(X_{(1)}>u)
=\prod_{i=1}^nP(X_i>u).
$$

For one shifted-exponential observation,

$$
P(X_i>u)
=\int_u^\infty \lambda e^{-\lambda(t-\theta)}\,dt
=e^{-\lambda(u-\theta)}.
$$

Therefore,

$$
P(X_{(1)}>u)
=\left[e^{-\lambda(u-\theta)}\right]^n
=e^{-n\lambda(u-\theta)}.
$$

Hence the CDF is

$$
F_{X_{(1)}}(u)
=P(X_{(1)}\le u)
=1-e^{-n\lambda(u-\theta)},
\qquad u\ge\theta.
$$

Differentiating,

$$
f_{X_{(1)}}(u)
=n\lambda e^{-n\lambda(u-\theta)},
\qquad u\ge\theta.
$$

Thus

$$
\boxed{
\hat \theta = X_{(1)}\sim \text{shifted exponential}(\theta,n\lambda).
}
$$

The maximum likelihood estimator of $\hat\theta$ has the same distributional family as the data distribution, but its rate is $n\lambda$ instead of $\lambda$.

```{r sampling-dsn-theta}
hist(
  theta_hat_sim,
  breaks = 50,
  probability = TRUE,
  main = expression("Sampling Distribution of " * hat(theta)),
  xlab = expression(hat(theta)),
  col = "skyblue",
  border = "white"
)

u_grid <- seq(theta_true, max(theta_hat_sim), length.out = 500)

exact_density <- n * lambda_true *
  exp(-n * lambda_true * (u_grid - theta_true))

lines(u_grid, exact_density, lwd = 2, col = "navy")

abline(v = theta_true, lty = 2)

legend(
  "topright",
  legend = c("Simulated", "Exact"),
  fill = c("skyblue", NA),
  border = c("white", NA),
  lty = c(NA, 1),
  lwd = c(NA, 2),
  bty = "n"
)
```

The histogram is empirical; the curve is the exact theoretical sampling distribution. With many repetitions (that's why we use a large `B`), they should agree closely.

Compare the theoretical results with the simulation.

```{r simulation-vs-theory}
comparison <- data.frame(
  quantity = c("Mean", "Variance", "Standard error"),
  simulation = c(
    mean(theta_hat_sim),
    var(theta_hat_sim),
    sd(theta_hat_sim)
  ),
  theory = c(
    theta_true + 1 / (n * lambda_true),
    1 / (n * lambda_true)^2,
    1 / (n * lambda_true)
  )
)

knitr::kable(
  comparison,
  digits = 4,
  caption = "Simulation versus exact sampling-distribution results"
)
```

# Bias and Variance

The bias of an estimator is its average distance from the true value of the population parameter:

$$
\mathrm{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta.
$$ We call an estimator **unbiased** if the bias is zero, that is, if $E(\hat{\theta}) = \theta$.

## Bias and variance of $\hat\theta$

$$
E(\widehat\theta)
=\theta+\frac{1}{n\lambda},
$$

and

$$
\operatorname{Var}(\widehat\theta)
=\frac{1}{(n\lambda)^2}.
$$Thus,

$$
\operatorname{Bias}(\widehat\theta)
=\frac{1}{n\lambda},
$$and

$$
\operatorname{SE}(\hat\theta) = \frac{1}{n\lambda}.
$$ $\hat\theta = X_{(1)}$ is a biased estimator of $\theta$, but the bias goes to $0$ as $n\to \infty$.

We also see that the standard error gets smaller with more data ($n$ is large).

## Estimate the bias and variance of $\hat\theta$

Since the theoretical formulas above involve the unknown rate $\lambda$, we can use a **plug-in estimate** that replaces $\lambda$ by $\widehat\lambda$:

$$
\widehat{\operatorname{Bias}}(\widehat\theta)=\frac1{n\widehat\lambda},
\qquad
\widehat{\operatorname{Var}}(\widehat\theta)=\frac1{n^2\widehat\lambda^2},
\qquad
\widehat{\operatorname{SE}}(\widehat\theta)=\frac1{n\widehat\lambda}.
$$ For $n=20$, $\theta=5$, and $\lambda=0.1$:

```{r plugin-estimates}
bias_hat <- 1 / (n * lambda_hat)
variance_hat <- 1 / (n * lambda_hat)^2
se_hat <- sqrt(variance_hat)
knitr::kable(
  data.frame(
    quantity = c("Bias", "Variance", "Standard error"),
    true_value = c(1 / (n * lambda_true), 1 / (n * lambda_true)^2,
                   1 / (n * lambda_true)),
    plugin_estimate = c(bias_hat, variance_hat, se_hat)
    ), 
  digits = 4
)
```

## What changes when the sample size increases?

The data distribution *does not* change when we collect a larger sample. However, the sampling distribution of $\widehat\theta$ *does* change because its rate is $n\lambda$.

Create a helper function that simulates many samples and calculate the estimate of $\theta$ for each sample.

```{r theta-sim-function}
sim_theta_hat <- function(n, B, theta, lambda) {
  replicate(
    B,
    min(theta + rexp(n, rate = lambda))
  )
}
```

Simulate three different sample sizes.

```{r bias-and-se-comparison-across-sample-size}
sample_sizes <- c(5, 20, 100)
theta_hat_by_n <- lapply(
  sample_sizes,
  sim_theta_hat,
  B = 10000,
  theta = theta_true,
  lambda = lambda_true
)
knitr::kable(
  data.frame(
    n = sample_sizes,
    simulated_bias = sapply(theta_hat_by_n, mean) - theta_true,
    theoretical_bias = 1 / (sample_sizes * lambda_true),
    simulated_SE = sapply(theta_hat_by_n, sd),
    theoretical_SE = 1 / (sample_sizes * lambda_true)
  ), 
  digits = 4
)
```

```{r sample-size-comparison, fig.height=4}
par(mfrow = c(1, 3))
for (j in seq_along(sample_sizes)) {
  nn <- sample_sizes[j]
  hist(theta_hat_by_n[[j]], probability = TRUE, breaks = 35,
       xlim = c(theta_true, theta_true + 10),
       col = "skyblue", border = "white", main = paste("n =", nn),
       xlab = expression(hat(theta)))
  curve(dexp(x - theta_true, rate = nn * lambda_true),
        add = TRUE, col = "navy", lwd = 2)
  abline(v = theta_true, col = "firebrick", lty = 2,  lwd = 2)
}
```

As $n$ grows, the sampling distribution becomes more concentrated around $\theta$. Both its bias and standard error decrease as the sample size increases.

# Sampling Distribution of $\widehat\lambda$

The exact sampling distribution of

$$
\widehat\lambda
=\frac{1}{\overline X-X_{(1)}}
$$

is more complicated than the sampling distribution of $\hat\theta = X_{(1)}$. It still has an exact sampling distribution though, it's just difficult to figure out. So we simulate its sampling distribution.

```{r lambda-sim-function}
sim_lambda_hat <- function(n, B, theta, lambda) {
  replicate(B, {
    x_b <- theta + rexp(n, rate = lambda)
    1 / (mean(x_b) - min(x_b))
  })
}
```

By doing simulations with several sample sizes, we can see that as the sample size increases, the simulated mean of $\hat\lambda$ gets closer to the true $\lambda$ and the simulated standard error of $\hat\lambda$ approaches zero.

```{r lambda-sampling-distributions}
sample_sizes <- c(5, 20, 100)

lambda_sim_by_n <- lapply(
  sample_sizes,
  sim_lambda_hat,
  B = 10000,
  theta = theta_true,
  lambda = lambda_true
)

lambda_summary <- data.frame(
  n = sample_sizes,
  simulated_mean = sapply(lambda_sim_by_n, mean),
  simulated_sd = sapply(lambda_sim_by_n, sd),
  true_lambda = lambda_true
)

knitr::kable(lambda_summary, digits = 4)
```

Let's plot the simulated sampling distribution of $\lambda$ for different sample sizes. For larger $n$, we can see that the simulated sampling distribution of $\lambda$ concentrate more tightly near the true $\lambda$. What do you find from the shape of the simulated sampling distribution?

```{r lambda-samping-dsn-plot, fig.height=4}
par(mfrow = c(1, 3))

for (j in seq_along(sample_sizes)) {

  nn <- sample_sizes[j]

  hist(
    lambda_sim_by_n[[j]],
    probability = TRUE,
    breaks = 60,
    col = "skyblue",
    border = "white",
    main = paste("n =", nn),
    xlab = expression(hat(lambda))
  )

  # Corresponding asymptotic normal density
  curve(
    dnorm(
      x,
      mean = lambda_true,
      sd = lambda_true / sqrt(nn)
    ),
    add = TRUE,
    lwd = 2,
    col = "navy"
  )

  # True value of lambda
  abline(v = lambda_true, lty = 2)
}

```
