---
title: "Confidence Intervals: Student Version"
author: "STAT 135 Lab 5"
date: "Fall 2026"
output: pdf_document
header-includes:
  - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
```

## 1. Components of a Confidence Interval

When working with real data, we essentially never know the true population variance ($\sigma^2$). Therefore, we focus on confidence intervals built from the data we actually have. The general form of a confidence interval for a sample mean $\bar{X}$ is:

$$\bar{X} \pm (\text{a quantile}) \times \widehat{SE}(\bar{X})$$

Here is how we find these pieces:

* **Point Estimate:** $\bar{X}$ (the sample mean).
* **Standard Error:** $\widehat{SE}(\bar{X}) = \frac{S}{\sqrt{n}}$, where $S$ is the sample standard deviation.
* **Quantile:** Because we estimate $\sigma$ with $S$, we introduce extra variability. For small samples, we use a quantile from the **$t$-distribution** with $n-1$ degrees of freedom instead of the standard normal distribution. (For large $n$, the $t$-distribution and normal distribution are virtually identical).

**Note on "Known Variance":** In lecture, creating confidence intervals with a known $\sigma^2$ was discussed. This is purely a thought experiment to build intuition. In real-world practice, we always have to estimate variance from the sample.

### True/False: Common Mistakes to Avoid
Test your understanding of the components above by answering True or False to the following statements:

1. **True / False:** To find the margin of error for a sample mean, you multiply the appropriate quantile by the sample standard deviation ($S$).
   
   \textcolor{violet}{\textbf{Answer:}} False. S/sqrt(n)

2. **True / False:** The standard error of the sample mean is calculated as $S^2/n$.
   
   \textcolor{violet}{\textbf{Answer:}} False. sqrt(S^2/n)

3. **True / False:** To compute a 95% confidence interval ($\alpha = 0.05$), you should use the 0.95 quantile of the appropriate distribution.
   
   \textcolor{violet}{\textbf{Answer:}} False. 0.975

4. **True / False:** A 95% confidence interval means that the true population parameter $\mu$ is a random variable that has a 95% chance of falling within our calculated boundaries.
   
   \textcolor{violet}{\textbf{Answer:}} False. $\mu$ is not a random variable, but a parameter.

## 2. Finding Quantiles in R

To build a $100(1-\alpha)\%$ confidence interval, you need the $1 - \alpha/2$ quantile.

**Why $1-\alpha/2$?**
If we want to be $95\%$ confident, our $\alpha = 0.05$ represents our total error rate. Because a confidence interval is symmetric, we split this error equally into the left and right tails (0.025 in each). To capture the middle $95\%$, we need the value that leaves exactly $2.5\%$ in the right tail, which is the $1 - 0.025 = 0.975$ cumulative probability quantile.

**Exercise:** What quantiles would you need for a 90% CI and a 99% CI? Fill in the missing values below.

```{r}
n <- 40

# Find the t-quantiles for 90% and 99% CIs:
t_90 <- qt(0.95, df = 39)
t_99 <- qt(0.995, df = 39)

cat("90% t-quantile:", round(t_90, 4), "\n")
cat("99% t-quantile:", round(t_99, 4), "\n")
```

**Conceptual Questions:**

* What happens to the width of the confidence interval if we increase our confidence level from 90% to 99%?
  
  \textcolor{violet}{\textbf{Answer:}} Wider.

* What happens to the $t$-quantile as the degrees of freedom ($n-1$) increases?
  
  \textcolor{violet}{\textbf{Answer:}} Smaller.

## 3. Example: Chicago Income Data (Mean)

Let's look at the Chicago public housing income data (from Agresti 4.9). We have a sample of 40 households.

Notice that to calculate the confidence interval, you don't actually need the raw dataset. You *only* need the necessary summary statistics. However, to check **assumptions** (like normality), you *do* need the full dataset.

```{r load-chicago}
chicago <- read.table("Chicago.dat", header = TRUE)

# Summary statistics table with extra distractor metrics
stats_table <- data.frame(
  Statistic = c(
    "Sample Size (n)",
    "Sample Mean (X-bar)",
    "Sample Variance (S^2)",
    "Sample Standard Deviation (S)",
    "Median",
    "Interquartile Range (IQR)",
    "Minimum",
    "Maximum",
    "Sum (sum X_i)"
  ),
  Value = c(
    nrow(chicago),
    round(mean(chicago$income), 3),
    round(var(chicago$income), 3),
    round(sd(chicago$income), 3),
    round(median(chicago$income), 3),
    round(IQR(chicago$income), 3),
    round(min(chicago$income), 3),
    round(max(chicago$income), 3),
    round(sum(chicago$income), 3)
  )
)

knitr::kable(stats_table, caption = "Summary Statistics: Chicago Household Income (in $1,000s)")
```

**Task:** Identify the necessary summary statistics from the table above, and calculate a 90% and 99% confidence interval for the true mean income.

```{r ci-chicago}
# Identify and extract only the relevant statistics from the table:
n_chic    <- nrow(chicago)
mean_chic <- mean(chicago$income)
sd_chic   <- sd(chicago$income)

# Calculate the Standard Error:
se_chic <- sd_chic / sqrt(n_chic)

# 90% Confidence Interval:
margin_90 <- qt(0.95, df = 29) * se_chic
ci_90     <- c(mean_chic - margin_90, mean_chic + margin_90)
cat("90% CI: (", round(ci_90[1], 3), ",", round(ci_90[2], 3), ")\n")

# 99% Confidence Interval:
margin_99 <- qt(0.995, df = 29) * se_chic
ci_99     <- c(mean_chic - margin_99, mean_chic + margin_99)
cat("99% CI: (", round(ci_99[1], 3), ",", round(ci_99[2], 3), ")\n")
```

We plot a histogram of the data below to inspect assumptions:

```{r chicago-hist, fig.height=3.5, fig.width=5.5, fig.align='center'}
hist(
  chicago$income,
  breaks = 10,
  col = "lightblue",
  border = "white",
  main = "Distribution of Household Income in Public Housing",
  xlab = "Income (in Thousands of Dollars)",
  ylab = "Frequency"
)
```

*Question:* If we check the histogram of this data, it's heavily right-skewed. Does the normality assumption hold? Why is the $t$-distribution still the safer choice here over the normal distribution?

\textcolor{violet}{\textbf{Answer:}} Due to the uncertainty, it is safer to use the $t$-distribution, which is more conservative.


## 4. Example: Belief in Life After Death (Proportions)

For proportions, the Central Limit Theorem allows us to use the normal approximation (Wald interval) if the sample size is large enough. The standard error for a proportion is:

$$\widehat{SE}(\widehat{\pi}) = \sqrt{\frac{\widehat{\pi}(1-\widehat{\pi})}{n}}$$

*(Note: The textbook also discusses forming CIs for proportions by rederiving them directly from the pivotal statistic, known as the Score interval. We are using the standard Wald interval here).*

In a 2018 survey, 1720 out of 2123 people said they believed in life after death.

**Task:** Calculate the 90% and 99% confidence intervals for the true population proportion.

```{r ci-proportion}
n_prop <- 2123
x_prop <- 1720

# Point estimate
pi_hat <- 1720 / 2123

# Calculate Standard Error
se_prop <- sqrt(pi_hat * (1 - pi_hat) / n_prop)

# Normal quantiles
z_90 <- qnorm(0.95)
z_99 <- qnorm(0.995)

# 90% CI: vector c(lower, upper)
ci_90 <- c(pi_hat - (z_90 * se_prop), pi_hat + (z_90 * se_prop))
cat("90% CI: (", round(ci_90[1], 4), ",", round(ci_90[2], 4), ")\n")

# 99% CI: vector c(lower, upper)
ci_99 <- c(pi_hat - (z_99 * se_prop), pi_hat + (z_99 * se_prop))
cat("99% CI: (", round(ci_99[1], 4), ",", round(ci_99[2], 4), ")\n")
```

**Compare with R:** Try running `prop.test(x = 1720, n = 2123, conf.level = 0.90)` in your console. Does the interval exactly match your hand calculation? Why might it be slightly different?

\textcolor{violet}{\textbf{Answer:}} No. It is because of the continuity correction.

## 5. Example: Two Groups (Difference in Means)

When comparing two independent groups, assuming equal variance between them is often unrealistic in practice. Let's look at an extreme example to see why.

* **Group 1:** $n_1 = 10, \quad \bar{X}_1 = 100, \quad S_1 = 5$
* **Group 2:** $n_2 = 40, \quad \bar{X}_2 = 90, \quad S_2 = 25$

Notice that Group 2 is much larger, but also has a much larger variance.

**Task:** Calculate the Standard Error for both the Equal Variance case and the Unequal Variance case.

$$\text{Equal Variance } SE = \sqrt{S_p^2 \left(\frac{1}{n_1} + \frac{1}{n_2}\right)} \quad \text{where} \quad S_p^2 = \frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1+n_2-2}$$

$$\text{Unequal Variance } SE = \sqrt{\frac{S_1^2}{n_1} + \frac{S_2^2}{n_2}}$$

```{r ci-two-groups}
n1 <- 10; xbar1 <- 100; s1 <- 5
n2 <- 40; xbar2 <- 90;  s2 <- 25

# Equal Variance SE
sp2      <- ((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2)
se_equal <- sqrt(sp2 * (1/n1 + 1/n2))
cat("Equal Variance SE:", round(se_equal, 3), "\n")

# Unequal Variance SE
se_unequal <- sqrt(s1^2/n1 + s2^2/n2)
cat("Unequal Variance SE:", round(se_unequal, 3), "\n")
```

Notice that the two standard errors are quite different ($8.004$ vs. $4.257$). When we assume equal variance, the much larger variance of Group 2 inflates the pooled estimate, which artificially increases the standard error for Group 1 and makes the resulting confidence interval nearly twice as wide. In practice, we rarely have a good reason to assume equal variances, so using the unequal variance (Welch) approach is standard.

For the unequal variance degrees of freedom, the formula (Welch's approximation) is complex, but for hand calculations, you can safely use the lower bound: $\min(n_1-1, n_2-1)$. In this case, we would use $10 - 1 = 9$ degrees of freedom.

**Task:** Using the standard errors calculated above, compute the 95% confidence intervals for both the pooled and Welch methods and compare their boundaries:

```{r ci-two-groups-intervals}
diff_mean <- xbar1 - xbar2

# Equal variance CI: find t-quantile and construct interval vector c(lower, upper)
t_pool  <- qt(0.975, df = n1 + n2 - 2)
ci_pool <- c(diff_mean - t_pool * se_equal, diff_mean + t_pool * se_equal)


# Unequal variance CI: find conservative t-quantile and construct interval vector
t_welch  <- qt(0.975, df = min(n1 - 1, n2 - 1))
ci_welch <- c(diff_mean - t_welch * se_unequal, diff_mean + t_welch * se_unequal)

cat("Equal Variance 95% CI:   (", round(ci_pool[1], 2), ",", round(ci_pool[2], 2), ")\n")
cat("Unequal Variance 95% CI: (", round(ci_welch[1], 2), ",", round(ci_welch[2], 2), ")\n")
```

**Question:** Is assuming equal variance merely a neutral mathematical convenience, or can it impact your conclusions?

\textcolor{violet}{\textbf{Answer:}} Assuming equal variance is not merely a neutral mathematical convenience. If the variances are actually different, the equal-variance assumption can substantially change the standard error and confidence interval, potentially changing our statistical conclusion.

## 6. Visualizing Confidence Intervals (Simulation)

To really understand what "95% Confidence" means, let's simulate 100 different samples from a known population and calculate a 95% confidence interval for each one.

We will draw samples of size $n=30$ from a Normal distribution with a true mean of $\mu = 50$ and true standard deviation $\sigma = 10$.

*(Remove `eval=FALSE` from the chunk header when you are ready to run the code and plot the intervals).*

```{r ci-sim, warning=FALSE, message=FALSE, fig.height=3.8, fig.width=6, fig.align='center'}
set.seed(135) # Set seed for reproducibility

n_sim <- 100
n_samp <- 30
true_mu <- 50
true_sigma <- 10

results <- data.frame(
  sim_id = 1:n_sim,
  sample_mean = numeric(n_sim),
  lower_ci = numeric(n_sim),
  upper_ci = numeric(n_sim)
)

for (i in 1:n_sim) {
  # Draw a sample
  samp <- rnorm(n_samp, mean = true_mu, sd = true_sigma)
  
  # Calculate statistics
  results$sample_mean[i] <- mean(samp)
  se <- sd(samp) / sqrt(n_samp)
  margin <- qt(0.975, df = n_samp - 1) * se
  
  # Store bounds
  results$lower_ci[i] <- results$sample_mean[i] - margin
  results$upper_ci[i] <- results$sample_mean[i] + margin
}

# Task: Create a logical column that is TRUE if the interval covers true_mu,
# and FALSE if it does not.
results$covers <- results$lower_ci <= true_mu & results$upper_ci >= true_mu

# Plot the 100 intervals
ggplot(results, aes(x = sample_mean, y = sim_id, color = covers)) +
  geom_point(size = 1.2) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0, linewidth = 0.6) +
  geom_vline(xintercept = true_mu, linetype = "dashed", color = "black", linewidth = 0.8) +
  theme_minimal() +
  labs(
    title = "100 Simulated 95% Confidence Intervals",
    subtitle = "How many intervals actually capture the true mean?",
    x = "Sample Mean",
    y = "Simulation Number",
    color = "Covers True Mean?"
  ) +
  scale_color_manual(values = c("FALSE" = "firebrick", "TRUE" = "steelblue"))
```

```{r ci-sim-summary}
n_misses <- sum(!results$covers)
cat("Observed misses:", n_misses, "out of", n_sim, "\n")
cat("Observed coverage rate:", mean(results$covers), "\n")
```

**Discussion Note: Why aren't there exactly 5 intervals missing the true mean?**

* Each confidence interval construction is an independent trial with a $0.05$ probability of failing to cover $\mu$.
* If we define $Y$ as the total number of non-covering intervals out of $100$, then $Y \sim \text{Binomial}(n = 100, p = 0.05)$.
* The expected number of misses is $E(Y) = 100 \times 0.05 = 5$.

* **Why it's not always 5:**
  
  \textcolor{violet}{\textbf{Explanation:}} The expected value of 5 is a long-run average, not a guarantee for each set of 100 confidence intervals.

* **Probability of exactly 5:**
  
  \textcolor{violet}{\textbf{Explanation:}} Using the binomial distribution, there is only about an 18% probability of observing exactly 5 misses, even though the expected number of misses is 5.

## 7. Appendix: Estimating Variance - Why $S^2$ is Unbiased vs $1/n$

Let us find an **unbiased** estimator for the population variance $\sigma^2$. First, look at the expected value of the sum of squared deviations from the sample mean, and expand it by adding and subtracting the true population mean $\mu$:

$$\sum_{i=1}^n (X_i - \bar{X})^2 = \sum_{i=1}^n ((X_i - \mu) - (\bar{X} - \mu))^2$$

**(Complete on the board)**

\vspace{4cm}

**Conclusion:**
If we used the intuitive $\frac{1}{n}$ formula to estimate variance, the expected value would be:
$$E\left[\frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2\right] = \frac{n-1}{n}\sigma^2$$
This means dividing by $n$ creates a **biased** estimator that systematically underestimates the true population variance.

However, if we define our sample variance $S^2$ by dividing by $n-1$, we perfectly cancel out the $(n-1)$ term:
$$E\left[\frac{1}{n-1}\sum_{i=1}^n (X_i - \bar{X})^2\right] = \frac{n-1}{n-1}\sigma^2 = \sigma^2$$
This is why $S^2$ is an **unbiased** estimator of $\sigma^2$.

