---
title: "Confidence Intervals and MLE Asymptotics"
subtitle: "Student Version"
author: "STAT 135 Lab 6"
date: "Fall 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.align = "center"
)
set.seed(13506)
```

<!-- #region -->
This lab connects three settings: paired measurements, a regression
slope, and a one-parameter MLE. In each, we assemble a confidence interval
from an estimate, an estimated standard error, and a critical value.
We will use a 95% confidence level unless a question specifies otherwise.

> Note that we use R **as a calculator**: we build the calculations from `mean()`, `sd()`,
`sum()`, and `qt()`, then use `t.test()` and `lm()` to verify the results. This is what you are expected to do in homeworks. 


## 1. Paired Data: Before and After Measurements

Rice, Section 11.3.1, page 447, reports platelet aggregation for 11 people
before and after smoking a cigarette. The measurements are percentages
of platelets that aggregated following a stimulus. 
<!-- #endregion -->

```{r paired-data, fig.width=5, fig.height=5}
paired <- read.csv("rice_p447_paired.csv")

limits <- range(paired$before, paired$after)
plot(paired$before, paired$after, pch = 19, col = "steelblue",
     xlim = limits, ylim = limits, asp = 1,
     xlab = "Before (%)", ylab = "After (%)",
     main = "Platelet Aggregation: Matched Measurements")
abline(a = 0, b = 1, lty = 2, col = "gray40")
```

### Understand the pairs

The same person supplies the before and after measurements. We therefore
keep each pair together throughout the analysis. The dashed line represents
no change: a point above it has a larger after measurement. The plot shows
a strong positive association between the two measurements.

**Quick checkpoint 1:** True or false: 11 people measured twice give us 22 independent observations.

### Organizing the measurements

The supplied file is in **wide format**, with before and after in separate
columns. In **long format**, the same data have one measurement per row,
a condition column, and a person ID. Preserving the ID is essential.

```{r paired-long-format}
paired_long <- reshape(paired,
                       varying = c("before", "after"),
                       v.names = "aggregation",
                       timevar = "condition",
                       times = c("before", "after"),
                       idvar = "id", direction = "long")
head(paired_long)
```

Define $D_i=\text{after}_i-\text{before}_i$ and let
$\mu_D=\mathrm{E}(D_i)$. Under independent, identically distributed normal
differences, the interval for $\mu_D$ is
$$
\bar D\ \pm\ t_{n-1,\,0.975}\frac{S_D}{\sqrt n},
$$
where $S_D$ is the sample standard deviation of the differences and
$t_{\nu,\,0.975}$ is the 0.975 quantile of a $t$ distribution with $\nu$
degrees of freedom. R computes it with `qt(0.975, df = ...)`.

### Calculate the interval step by step

We calculate the changes first, then treat them as a single sample.
`sd(d)` measures variation among individual changes. Dividing by
$\sqrt n$ estimates the standard error of their average. We assemble
the margin of error from the $t$ critical value and this SE.

```{r walkthrough-2}
d <- paired$after - paired$before
n_pair <- length(d)
d_bar <- mean(d)
sd_d <- sd(d)
se_d <- sd_d / sqrt(n_pair)
t_star <- qt(0.975, df = n_pair - 1)
ci_paired <- d_bar + c(-1, 1) * t_star * se_d
c(mean_change = d_bar, SD = sd_d, SE = se_d,
  critical_value = t_star, lower = ci_paired[1], upper = ci_paired[2])
```

The average increase is about 10.273 percentage points. The estimated
SE is 2.405, giving the 95% interval $(4.914,15.631)$ percentage points
for the population mean change. As we have talked about in the lasst his is not an interval containing
95% of individual changes: confidence refers to repeated-sample
coverage of the fixed population mean.

The exact $t$ result assumes independent, identically distributed
normal **differences**. With only 11 pairs, strong skewness or outliers
in the differences would matter; choosing a $t$ quantile does not
remove that concern.

**Quick checkpoint 2:** For a 99% two-sided interval, would we use `qt(0.99, ...)` or `qt(0.995, ...)`?

### Compare with an analysis that ignores pairing

For comparison only, pretend these measurements came from two
independent groups. We use the same difference of means but estimate
its SE without a covariance term. Let $A$ denote after and $B$ before:
$$
a=\frac{s_A^2}{n_A},\qquad b=\frac{s_B^2}{n_B},\qquad
\widehat{\mathrm{SE}}_{\mathrm{ind}}=\sqrt{a+b}.
$$
For the Welch interval, use the approximate degrees of freedom
$$
\nu=\frac{(a+b)^2}{a^2/(n_A-1)+b^2/(n_B-1)} 
\geq \min\{n_A - 1, n_B - 1\}. 
$$
The following is a deliberately incorrect independence analysis of
these paired data, included to show why recognizing the design matters.

```{r walkthrough-3}
n_after <- length(paired$after)
n_before <- length(paired$before)
a <- var(paired$after) / n_after
b <- var(paired$before) / n_before
se_ind <- sqrt(a + b)
df_ind <- (a + b)^2 / (a^2 / (n_after - 1) + b^2 / (n_before - 1))
ci_ind <- (mean(paired$after) - mean(paired$before)) +
  c(-1, 1) * qt(0.975, df = df_ind) * se_ind
knitr::kable(data.frame(
  calculation = c("Paired", "Ignoring pairing"),
  SE = c(se_d, se_ind),
  lower = c(ci_paired[1], ci_ind[1]),
  upper = c(ci_paired[2], ci_ind[2]),
  width = c(diff(ci_paired), diff(ci_ind))
), digits = 3)
```

```{r}
# Sample correlation calculation 
A <- paired$after
B <- paired$before

sample_cov <- sum((A - mean(A)) * (B - mean(B))) /
  (length(A) - 1)

r_AB <- sample_cov / (sd(A) * sd(B))
r_AB

cor(A, B)  # Same result
```

The point estimate is unchanged, but ignoring pairing gives a much
wider interval. The reason is the covariance term:
$$
\mathrm{Var}(A-B)=\mathrm{Var}(A)+\mathrm{Var}(B)
-2\mathrm{Cov}(A,B).
$$
Here the before-after correlation is about 0.901. Positive covariance
reduces the variance of a difference. With independent pairs, divide
that variance by the number of pairs to get the variance of the mean
change. Pairing removes persistent differences between people and can
improve precision. It is not universally variance-reducing: a negative
within-pair covariance would have the opposite effect.

> **Question** : Why does it make sense that a negative correlation increases the variance of the difference? 
    
### Check the result with R

Now that we have assembled the intervals ourselves, we can check them
with `t.test()`. The first argument minus the second determines the
direction of the difference. `paired = TRUE` tells R to use within-person
differences; `$conf.int` extracts just the interval, without test output.
The one-sample call on `d` should reproduce the paired calculation.

```{r walkthrough-4}
check_paired <- t.test(paired$after, paired$before,
                      paired = TRUE, conf.level = 0.95)$conf.int
check_d <- t.test(d, conf.level = 0.95)$conf.int
check_ind <- t.test(paired$after, paired$before,
                   paired = FALSE, conf.level = 0.95)$conf.int
rbind(paired = check_paired, differences = check_d,
      ignoring_pairing = check_ind)
```

R pairs observations by their positions in the two vectors. Keep
rows aligned by person ID; independently sorting the two columns
would destroy the matching even if `paired = TRUE` were specified.

**Quick checkpoint 3:** If we switch from after minus before to before minus after, does the interval's width change?

## 2. Regression: Living Area and Sale Price

We use 2,930 Ames, Iowa, home sales from the dataset documented by
De Cock (2011). The supplied CSV contains the two columns needed here,
with names matching `AmesHousing::make_ames()`. `Gr_Liv_Area` is
above-ground living area in square feet and `Sale_Price` is in dollars.

```{r ames-data-and-plots, fig.width=8, fig.height=4}
ames <- read.csv("ames_area_price.csv")
stopifnot(nrow(ames) == 2930, !anyNA(ames),
          all(ames$Gr_Liv_Area > 0), all(ames$Sale_Price > 0))
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(Sale_Price ~ Gr_Liv_Area, data = ames,
     pch = 19, cex = 0.35, col = "gray45",
     xlab = "Living area (sq ft)", ylab = "Sale price ($)",
     main = "Original Axes")
plot(Sale_Price ~ Gr_Liv_Area, data = ames, log = "xy",
     pch = 19, cex = 0.35, col = "gray45",
     xlab = "Living area (sq ft)", ylab = "Sale price ($)",
     main = "Logarithmic Axes")
par(mfrow = c(1, 1))
```

For inference, model the transformed variables
$$
x_i=\log(\text{living area}_i),\qquad
Y_i=\log(\text{sale price}_i),\qquad
Y_i=\beta_0+\beta_1x_i+\epsilon_i.
$$
Conditional on the predictors, iid $N(0,\sigma^2)$ errors give the exact
$t$ interval below. Independence, linearity of the conditional mean,
and common error variance are substantive model assumptions.

The calculator ingredients are
$$
S_{xx}=\sum_i(x_i-\bar x)^2,\qquad
S_{xy}=\sum_i(x_i-\bar x)(y_i-\bar y),
$$
$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}},\quad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x,\quad
e_i=y_i-(\widehat\beta_0+\widehat\beta_1x_i),
$$
$$
s^2=\frac{\sum_i e_i^2}{n-2},\qquad
\widehat{\mathrm{SE}}(\widehat\beta_1)=\sqrt{\frac{s^2}{S_{xx}}}.
$$
A 95% CI is $\widehat\beta_1\pm t_{n-2,\,0.975}
\widehat{\mathrm{SE}}(\widehat\beta_1)$.

### Calculate the regression interval step by step

We first create the logged variables explicitly. Setting `log="xy"`
in a plot changes the axes but does not transform the stored data for
a later model. Then we calculate the coefficients, fitted residuals,
residual variance, and slope SE in that order.

The variance divisor is $n-2$ because we estimated an intercept and a
slope. Notice that the slope SE depends on both residual variation and
predictor spread; `sd(y)/sqrt(n)` would estimate the SE of a response
mean, not a slope.

```{r walkthrough-5}
x <- log(ames$Gr_Liv_Area)
y <- log(ames$Sale_Price)
n_reg <- length(y)
Sxx <- sum((x - mean(x))^2)
Sxy <- sum((x - mean(x)) * (y - mean(y)))
b1 <- Sxy / Sxx
b0 <- mean(y) - b1 * mean(x)
residual <- y - (b0 + b1 * x)
s2 <- sum(residual^2) / (n_reg - 2)
se_b1 <- sqrt(s2 / Sxx)
ci_b1 <- b1 + c(-1, 1) * qt(0.975, df = n_reg - 2) * se_b1
c(intercept = b0, slope = b1, slope_SE = se_b1,
  lower = ci_b1[1], upper = ci_b1[2])
```

The estimated slope is about 0.908, its estimated SE is 0.0160,
and the 95% CI is approximately (0.876, 0.939).

**Quick checkpoint 4:** If the SE stays fixed, does moving from 95% to 99% confidence make the slope interval wider or narrower?

### Using the model-fitting interface

`lm(response ~ predictor, data = data_frame)` fits a linear mean model
with an intercept by default. The expression to the left of `~` is the
response; the expression to the right specifies the predictor.
`log()` can be written inside the formula.

For a fitted model named `fit`, `coef(fit)` extracts coefficients,
`summary(fit)$coefficients[, 1:2]` extracts estimates and SEs without
p-values, and `confint(fit, level = 0.95)` returns coefficient CIs.
The function's name is **`confint()`**, with no dot.

### Fit the same model with lm()

We now fit the same log-log model with `lm()`. The coefficient
estimates, slope SE, and slope interval should match our calculator
results. To draw the fitted straight line with `abline(fit)`, we plot
log-price against log-area, the coordinates used by the model.

```{r walkthrough-6}
fit <- lm(log(Sale_Price) ~ log(Gr_Liv_Area), data = ames)
coef(fit)
summary(fit)$coefficients[, 1:2]
ci_lm <- confint(fit, "log(Gr_Liv_Area)", level = 0.95)
ci_lm
plot(x, y, pch = 19, cex = 0.4, col = "gray60",
     xlab = "log(living area)", ylab = "log(sale price)",
     main = "Fitted Log-Log Regression")
abline(fit, col = "firebrick", lwd = 2)
```

**Quick checkpoint 5:** In `lm(log(Sale_Price) ~ log(Gr_Liv_Area))`, which variable is the response?

### Interpret the fit and check assumptions

The fitted slope is about 0.908: a 1% increase in area is associated
with approximately a 0.908% increase in fitted geometric-mean price.
The interval concerns this population slope, not the price of an
individual house. This observational association is not automatically
a causal effect of adding space.

The normal, constant-variance error assumptions also need attention.
A residual-versus-fitted plot can reveal curvature or changing spread;
a normal Q-Q plot helps inspect tail departures. These plots show
outliers and tail departures, so the model assumptions should not be
accepted automatically. Large sample size alone does not fix an SE
formula that ignores heteroskedasticity or dependence.

```{r walkthrough-7}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fitted(fit), resid(fit), pch = 19, cex = 0.4, col = "gray50",
     xlab = "Fitted log-price", ylab = "Residual",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2, col = "firebrick")
qqnorm(resid(fit), pch = 19, cex = 0.4, main = "Normal Q-Q Plot")
qqline(resid(fit), col = "firebrick", lwd = 2)
par(mfrow = c(1, 1))
```

## 3. MLE Asymptotics: The Geometric Model

For iid observations and a regular one-parameter model, define
$$
I(\theta)=-\mathrm E_\theta\left[
\frac{\partial^2}{\partial\theta^2}\log f(X;\theta)\right],
\qquad I_n(\theta)=nI(\theta).
$$
$I(\theta)$ is information in **one observation**; $I_n(\theta)$ is
information in the **whole sample**. The derivatives are of the
**log** density or PMF, and the expectation uses the same parameter.

For a true interior parameter $\theta_0$, under regularity conditions,
$$
\sqrt{nI(\theta_0)}(\widehat\theta_n-\theta_0)
\xrightarrow{D}N(0,1).
$$
Thus a large-sample approximation is
$$
\widehat\theta_n\ \dot\sim\ N\left(\theta_0,
\frac{1}{nI(\theta_0)}\right),\qquad
\widehat{\mathrm{SE}}(\widehat\theta_n)
=\frac{1}{\sqrt{I_n(\widehat\theta_n)}}.
$$
A Wald interval uses this plug-in SE and `qnorm(0.975)`. This is an
asymptotic statement, not an assertion of exact normality or exact
finite-sample variance. The regularity conditions include a correctly
specified, identifiable model and sufficient smoothness; a
parameter-independent support alone is not the entire justification.

### Read the information formula

Information describes how sharply the log-likelihood is curved on
average. Expected information has averaged out the sample randomness;
it is a function of the parameter, not the particular observations.
For iid data, information adds across observations.

For example, if $I(\theta_0)=5$ and $n=100$, then $I_n(\theta_0)=500$.
The approximate variance is $1/500=0.002$, while the approximate SE is
$1/\sqrt{500}\approx0.0447$. Keeping information, variance, and SE
separate prevents an extra or missing square root in a CI.

### What is one geometric observation?

An experiment consists of independent Bernoulli trials with constant
success probability $\pi\in(0,1)$, continued until the first success.
Let $X$ count **trials including that success**:
$$
f(x;\pi)=(1-\pi)^{x-1}\pi,\quad x=1,2,\ldots,\qquad
\mathrm E_\pi(X)=\frac1\pi.
$$
Our sample contains the counts from $n$ independent repetitions of this
experiment. In R, `rgeom()` instead counts **failures before** the first
success; check the convention when translating the model into code.

### Simulate geometric observations

For example, the trial sequences `0,0,1`; `1`; and `0,1` give
trials-to-success counts of 3, 1, and 2. Each count is one geometric
observation. To simulate this convention in R, add 1 to the
failures-to-success count produced by `rgeom()`.

```{r walkthrough-9}
set.seed(13506)
n_geom <- 100
p_true <- 0.3
x_geom <- rgeom(n_geom, prob = p_true) + 1
head(x_geom)
```

**Quick checkpoint 6:** If `rgeom()` returns 0, how many trials occurred before and including the first success?

### Derive the MLE and Fisher information

Independence gives
$$
\ell(\pi)=\pi^n(1-\pi)^{\sum_i x_i-n},\qquad
L(\pi)=n\log\pi+\left(\sum_i x_i-n\right)\log(1-\pi).
$$
The score is
$$
L'(\pi)=\frac n\pi-\frac{\sum_i x_i-n}{1-\pi}.
$$
Setting it to zero yields $n=\pi\sum_i x_i$, so
$\widehat\pi=n/\sum_i x_i=1/\bar x$.
For a single observation,
$$
\frac{\partial\log f}{\partial\pi}
=\frac1\pi-\frac{X-1}{1-\pi},\qquad
\frac{\partial^2\log f}{\partial\pi^2}
=-\frac1{\pi^2}-\frac{X-1}{(1-\pi)^2}.
$$
Hence
$$
I(\pi)=\frac1{\pi^2}+\frac{1/\pi-1}{(1-\pi)^2}
=\frac1{\pi^2(1-\pi)},\qquad
I_n(\pi)=\frac n{\pi^2(1-\pi)}.
$$
The full-sample derivative already includes all $n$ contributions.
Multiplying its expected negative second derivative by $n$ again would
incorrectly count information twice.

For samples containing at least one $x_i>1$, the MLE is interior.
If all $x_i=1$, the maximizer is the boundary $\pi=1$ when that boundary
is allowed; on the open parameter space there is only a supremum.
The interior Wald approximation does not apply to that fitted boundary.

### Estimate the standard error and form a Wald interval

For fixed $0<\pi_0<1$ and large $n$,
$$
\widehat\pi_n\ \dot\sim\ N\left(\pi_0,
\frac{\pi_0^2(1-\pi_0)}n\right),\qquad
\widehat{\mathrm{SE}}=\widehat\pi\sqrt{\frac{1-\widehat\pi}{n}}.
$$
Consistency and continuity justify plugging in the MLE:
$$
\frac{\sqrt n(\widehat\pi_n-\pi_0)}
{\widehat\pi_n\sqrt{1-\widehat\pi_n}}\xrightarrow{D}N(0,1).
$$
The interval is
$\widehat\pi\pm z_{0.975}\widehat\pi\sqrt{(1-\widehat\pi)/n}$.

The Fisher-information
variance is a large-sample approximation, not generally an exact
finite-sample variance of $1/\bar X$.

Small samples may give skewed sampling distributions and poor Wald
coverage. An all-ones sample gives a
boundary estimate (an estimate at the edge of the allowed parameter space) with a misleading zero plug-in SE. This makes the Wald interval collapse to a single point, which is a failure of the boudary approximation rather than evidence of perfect certainty. In addition, endpoints can leave $[0,1]$. In that case, simply clipping endpoints, i.e. intersecting the interval with [0,1], although removing the invalid values, does not establish nominal coverage. The code flags a boundary estimate rather than reporting that degenerate interval.

```{r walkthrough-11}
p_hat <- 1 / mean(x_geom)
if (p_hat <= 0 || p_hat >= 1) {
  stop("Boundary estimate: the interior Wald calculation is not appropriate.")
}
I_one_hat <- 1 / (p_hat^2 * (1 - p_hat))
I_sample_hat <- n_geom * I_one_hat
se_hat <- 1 / sqrt(I_sample_hat)
ci_geom <- p_hat + c(-1, 1) * qnorm(0.975) * se_hat
c(MLE = p_hat, information_one = I_one_hat,
  information_sample = I_sample_hat, estimated_SE = se_hat,
  lower = ci_geom[1], upper = ci_geom[2])
```

**Quick checkpoint 7:** True or false: `1 / I_sample_hat` is the estimated standard error.

### Visualize the sampling distribution

We now repeat the entire sampling-and-estimation process 2,000 times.
Each histogram entry is an MLE from one sample of 100 geometric counts.
The individual counts are discrete and right-skewed, but the distribution
of their MLE becomes approximately normal as the sample size grows.

Because this is a simulation, we know $\pi_0=0.3$ and can compare the
empirical SD with $0.3\sqrt{0.7/100}\approx0.0251$. In real data the
true parameter is unknown, so we use a plug-in SE instead.

```{r walkthrough-12}
set.seed(135061)
B <- 2000
estimates <- replicate(B, {
  draw <- rgeom(n_geom, prob = p_true) + 1
  1 / mean(draw)
})
asym_sd <- p_true * sqrt((1 - p_true) / n_geom)
c(empirical_mean = mean(estimates), truth = p_true,
  empirical_SD = sd(estimates), asymptotic_SD = asym_sd)
hist(estimates, probability = TRUE, breaks = 30,
     col = "gray85", border = "white",
     xlab = "MLE from one sample", main = "Sampling Distribution of the MLE")
curve(dnorm(x, mean = p_true, sd = asym_sd),
      add = TRUE, col = "firebrick", lwd = 2)
abline(v = p_true, lty = 2, col = "navy")
legend("topright", legend = c("Normal approximation", "True probability"),
       col = c("firebrick", "navy"), lty = c(1, 2), lwd = 2, bty = "n")
```

**Quick checkpoint 8:** If we quadruple the sample size, approximately what happens to the spread of the MLE histogram?

## 4. Attendance Question: Poisson Information

Suppose $X_1,\ldots,X_n$ are iid Poisson$(\lambda)$, with $\lambda>0$:
$$
f(x;\lambda)=\frac{\lambda^xe^{-\lambda}}{x!},
\quad x=0,1,2,\ldots.
$$
Recall that $\mathrm E(X)=\mathrm{Var}(X)=\lambda$ and the MLE is
$\widehat\lambda=\bar X$.

**Attendance checkpoint:** For the Poisson model, what is $I_n(\lambda)$?
What variance approximation does its reciprocal give for the MLE?
Is that variance also exact in this example?

